\documentclass{article}

\usepackage{graphicx}
\usepackage{url}
\usepackage{listings}
\usepackage{listings-rust}
\usepackage{makecell}
\usepackage{longtable}
\usepackage{multirow}
\usepackage{xcolor}
\usepackage{float}
\usepackage[a4paper, left=1.45cm, right=1.45cm, top=2cm, bottom=2cm]{geometry}
\usepackage{array}
\usepackage[hidelinks]{hyperref}

\definecolor{codegreen}{rgb}{0,0.6,0}
\definecolor{codegray}{rgb}{0.5,0.5,0.5}
\definecolor{codepurple}{rgb}{0.58,0,0.82}
\definecolor{backcolour}{rgb}{0.95,0.95,0.92}

% Prevent bibliography creating its own 'References' heading
\renewcommand{\refname}{\vspace*{-1\baselineskip}}

\lstdefinestyle{mystyle}{
  backgroundcolor=\color{backcolour},
  commentstyle=\color{gray},
  keywordstyle=\color{blue},
  numberstyle=\tiny\color{codegray},
  stringstyle=\color{orange},
  basicstyle=\ttfamily\footnotesize,
  breakatwhitespace=false,
  breaklines=true,
  captionpos=b,
  keepspaces=true,
  numbers=left,
  numbersep=5pt,
  showspaces=false,
  showstringspaces=false,
  showtabs=false,
  tabsize=2
}

\lstset{
  style=mystyle,
  frame=single
}

% subsubsubsection
\setcounter{tocdepth}{4}
\newcounter{subsubsubsection}[subsubsection]
\renewcommand\thesubsubsubsection{\thesubsubsection.\arabic{subsubsubsection}}
\newcommand{\subsubsubsection}[1]{%
  \refstepcounter{subsubsubsection}%
  \paragraph*{\thesubsubsubsection\enspace #1}%
  \addcontentsline{toc}{paragraph}{\protect\numberline{\thesubsubsubsection}#1}%
  \mbox{}\par
  \medskip
}

\newcommand{\layout}[1]{\parbox{\linewidth}{#1}}

\renewcommand{\arraystretch}{1.2}
\setlength{\parindent}{0em}

\newcommand{\myref}[1]{[\nameref{#1} \ref{#1}]}

% TODO: Add PIE (position independent executable)
% TODO: BIOS, UEFI, Multiboot spec
% TODO: ACCESS RIGHTS
% TODO: hidden instructions, appendix H, intels FDIV off by a small amount, 66 jmp (amd makes it 16 bit, intel ignores the 66)
% TODO: syntaxes? (intel vs at&t)

\title{\textbf{x86 Architecture}}
\author{William Bland}

\begin{document}

\maketitle
\tableofcontents
\newpage

\nocite{*}

\layout{
  \section{Introduction}

  This document contains the basics of the x86 architecture, and could be used as an introduction into the x86 architecture.
  It assumes you have some knowledge of programming and computer science.
  Additionally it does summarise some sections (such as instructions) to keep them concise, however it should sufficient to write basic x86 assembly programs (some examples can be seen in the appendix) or even a simple bare-metal program/kernel (can also be seen in the appendix);
  all code written is assembled with NASM and linked with ld.
  Furthermore, this document is mainly based off "\texttt{Intel® 64 and IA-32 Architectures Software Developer’s Manuals}", hence may have some content specific to the newest Intel CPUs.
}

\layout{
  \section{x86 architecture}

  Processors run by sequentially executing binary instructions (machine code),
  assembly code is a human readable version of machine code which can be converted (assembled) into machine code using an assembler.
  Every processor has an architecture, this architecture describes how the processor executes machine code (therefore different architectures execute machine code differently, so have different machine code and assembly languages).
  For multiple processors to be able to execute the same machine code they must have the same architecture, some popular architectures include ARM, PowerPC, RISC-V and x86.
  x86 is an example of a Complex Instruction Set Computer (CISC), so it uses a large set of instructions that can perform multiple operations in a single command.
  This is in opposition to Reduced Instruction Set Computer (RISC) which uses a smaller set of simple instructions.
  CISC (x86) is often more powerful than RISC due to its larger instruction set (allowing what would have been multiple RISC instructions to be ran as a single instruction), however this uses a lot more power therefore CISC (x86) is mainly used for desktop computers while RISC is mainly used for embedded and mobile devices.
}

\layout{
  \subsection{History of x86}

  Below is a list of a few of the notable x86 processors released by Intel.
  Furthermore, during the development of these processors and more recent processors, there were also significant improvements such as instruction pipelining, performance and features added such as SIMD and virtualisation.

  \begin{itemize}
    \item The first x86 processors were the 8086 and 8088 which were 16-bit.
      Through the use of segmentation, it allowed a 20-bit address bus so was able to address $2^{20}$ Bytes of memory (1MB).
    \item The Intel 286 processor introduced protected mode by uses the segment registers as selectors into descriptor tables which describe how a segment can be used (so the OS can ensure each process only accesses memory it has the permission to).
      Furthermore the descriptors provide a 24-bit base address so can address up to 16MB of memory.
    \item The Intel 386 processor was the first 32-bit processor, introducing 32-bit registers and paging.
    \item The Intel 486 introduced an integrated x87 FPU (previously FPUs were separate components).
    \item The Intel Pentium 4 introduced 64-bit extensions.
  \end{itemize}

  \cite[Chapter~2.1]{sdm1}
}

%\subsection{IBM compatible}
% TODO: BIOS + UEFI
% BIOS - enable A20

%\subsection{Branch prediction}
% TODO: Spectre/meltdown, pipelining

\layout{
  \subsection{Operating modes}

  x86 processors have several different operating modes that they can run in, these operating modes can change the addressing mode, available instruction set and registers.

  \begin{itemize}
    \item \textbf{Real Mode\cite{mode-real}}: \\
      Also called real address mode, all addresses always correspond to physical (real) memory addresses; memory addresses are 20 bit allowing for 1MiB of memory, however there is no memory protection.
      And all x86 CPU's start in real mode for backwards compatibility and start executing from address FFFF:0000
    \item \textbf{Protected Mode\cite{mode-protected}}: \\
      Also called protected virtual address mode, it allows the use of segmentation, virtual memory, paging and safe multi-tasking.
      To enter protected mode the software must setup a segment descriptor table and enable the Protection Enable (PE) bit in CR0
    \item \textbf{Long Mode\cite{mode-long}}: \\
      The 64-bit OS can access 64-bit registers and instructions.
      64-bit programs run in a sub-mode called 64-bit mode, and 32-bit and 16-bit protected mode programs run in another sub-mode called compatibility mode, however real mode and virtual 8086 mode programs cannot run natively in long mode
    \item \textbf{Unreal Mode\cite{mode-unreal}}: \\
      A variant of real mode which allows access to 4GiB of memory
    \item \textbf{System Management Mode\cite{mode-smm}}: \\
      Used by firmware (BIOS/UEFI) for power and hardware control.
      It is entered through a SMI (System Management Interrupt), and executes code in an address space isolated from other operating modes
    \item \textbf{Virtual 8086 Mode\cite{mode-virt8086}}: \\
      Also called virtual real mode, it allows real mode processes to run in protected mode. It is a hardware virtualisation technique and uses the 20-bit segmentation that real mode uses but is subject to the protected mode's memory paging
  \end{itemize}
}

\subsection{Multiprocessor}

In multiprocessor (multithreaded) environments, the processor is able to execute processes in parallel.
This can lead to race conditions if not handled correctly.

% Store buffer:

x86 processors store writes to memory in a store buffer.
It allows the processor to continue execution without having to wait for a write to cache/main memory to complete.
The store buffer is flushed when: a LOCK or a MFENCE/SFENCE is used, an I/O instruction executes or an exception/interrupt is generated \cite[Chapter~13.10]{sdm3}.

% TODO: APIC (ID)

% TODO: reorder buffer + out of order execution + multithreading + locks + mutex + synchronisation + visibility

% TODO: Virtualisation

\layout{
  \section{Memory}

  Processors operate on digital signals which can either be high or low (interpreted in binary as a 1 or a 0), a single 1 or 0 is a bit.
} \bigskip

\layout{
  Data sizes:
  \begin{table}[H]
    \centering
    \begin{tabular}{|c|c|c|c|}
      \hline \textbf{Name} & \textbf{Unit} & \textbf{Bits} & \textbf{Bytes} \\ \hline
      bit & b & 1 & 1/8 \\
      nibble & - & 4 & 1/2 \\
      byte & B & 8 & 1 \\
      word & - & 16 & 2 \\
      dword (double word) & - & 32 & 4 \\
      qword (quad word) & - & 64 & 8 \\ \hline
    \end{tabular}
  \end{table}
}

\layout{
  All memory addresses point to a unique byte in memory therefore a multiple byte integer (eg 64-bit) could be stored in two different orderings (its endianness). In Little-Endian (LE) ordering the least significant byte occupies the smaller address space, while in Big-Endian (BE) ordering the most significant byte occupies the smaller address space. x86 and most other processors use LE while data transmitted over networks uses BE.

  \bigskip \cite[Chapter~1.3.1]{sdm1}
}

\layout{
  \subsection{Registers}

  Resisters store a small amount of memory directly accessible to instructions.
}

\layout{
  \subsubsection{General Purpose Registers}

  General purpose registers are the most commonly accessed registers, they are used for a variety of uses such as arithmetic, logic and memory addressing.
  Most instructions require a register to be specified while some may implicitly use a specific register (such as stack operations modifying the stack pointer register).
  The registers on the same line in the table overlap in memory (eg the 32-bit register is located in the lower half of the 64-bit register).
  Furthermore both the lower and higher 8-bit registers exist within the 16-bit registers.
}

\begin{table}[H]
  \centering
  \begin{tabular}{|c|c|c|c|c|l|}
    \hline \textbf{64-bit} & \textbf{32-bit} & \textbf{16-bit} & \textbf{Higher 8-bit} & \textbf{Lower 8-bit} & \textbf{Description} \\ \hline
    RAX & EAX & AX & AH & AL & Accumulator \\
    RBX & EBX & BX & BH & BL & Base \\
    RCX & ECX & CX & CH & CL & Counter \\
    RDX & EDX & DX & DH & DL & Data (often extends the accumulator) \\
    RSI & ESI & SI & - & SIL & Source index for string operations \\
    RDI & EDI & DI & - & DIL & Destination index for string operations \\
    RSP & ESP & SP & - & SPL & Stack pointer \\
    RBP & EBP & BP & - & BPL & Base pointer (for stack frames) \\
    R8 & R8D & R8W & - & R8B & General purpose register \\
    R9 & R9D & R9W & - & R9B & General purpose register \\
    R10 & R10D & R10W & - & R10B & General purpose register \\
    R11 & R11D & R11W & - & R11B & General purpose register \\
    R12 & R12D & R12W & - & R12B & General purpose register \\
    R13 & R13D & R13W & - & R13B & General purpose register \\
    R14 & R14D & R14W & - & R14B & General purpose register \\
    R15 & R15D & R15W & - & R15B & General purpose register \\ \hline
  \end{tabular}
\end{table}

\layout{
  Registers 8-15, SIL, DIL, SPL and BPL are only available in 64-bit code since they were added in the 64-bit extensions.
  Furthermore, 64-bit code cannot access a higher 8-bit register and a register added in the 64-bit extensions in the same instruction.

  \bigskip \cite[Chapter~3.4.1]{sdm1}
}

% TODO: Add Instruction Pointer

\layout{
  \subsubsection{Segment Registers}

  Segment registers are 16-bit registers which are used for memory segmentation.
  The segment registers are:

  \begin{table}[H]
    \centering
    \begin{tabular}{|c|l|}
      \hline \textbf{Register} & \textbf{Description} \\ \hline
      CS & Code segment \\
      DS & Data segment \\
      SS & Stack segment \\
      ES & Extra segment (for string operations) \\
      FS & General purpose segment (Since i386) \\
      GS & General purpose segment (Since i386) \\ \hline
    \end{tabular}
  \end{table}
}

\layout{
  Furthermore, each segment register contains a hidden 64-bit register which is used as a segment descriptor cache.

  \bigskip \cite[Chapter~3.4.3]{sdm3}
}

% TODO: add gdtr/ldtr in the segment registers

\layout{
  \subsubsection{Control Registers}

  Control registers contain flags that control the operation of the CPU (such as enabling specific features).
  Control registers not listed in the table below such as CR1 and bits not listed are reserved and should be 0 (writing to a reserved register results in a \#UD exception, setting a reserved bit in CR0[0:31] is ignored, setting a reserved bit in CR0[32:63], CR4 or CR3[MAXPHYADDR:63] causes a \#GP).
  When protection is enabled, control registers can only be written to at CPL 0, however CR2 is writable by software.
  On a 64-bit processor, writing to a control register outside of 64-bit mode sets the upper 32 bits to 0, however when transitioning out of IA-32e mode, the registers retain their upper 32 bits.
  CR8 is only available in 64-bit mode, and CR4[32:63] only affects IA-32e mode and can only be changed in IA-32e mode.
} \bigskip

The following table is a list of the control registers and their flags:
\begin{longtable}{|c|c|m{0.65\linewidth}|}
  \hline \textbf{Bits} & \textbf{Label} & \textbf{Description} \\
  \hline \multicolumn{3}{|l|}{\textbf{CR0} - Contains system control flags that control operating mode and states of the processor} \\ \hline
  0 & PE & Protected Mode Enable \\
  1 & MP & Monitor co-processor \\
  2 & EM & x87 FPU Emulation \\
  3 & TS & Task switched \\
  4 & ET & Extension type \\
  5 & NE & Numeric error \\
  16 & WP & Write protect \\
  18 & AM & Alignment mask \\
  29 & NW & Not-write through \\
  30 & CD & Cache disable \\
  31 & PG & Paging \\
  \hline \multicolumn{3}{|l|}{\textbf{CR2} - Contains linear address that caused a page fault} \\ \hline
  0:31(63) & PFLA & Page Fault Linear Address \\
  \hline \multicolumn{3}{|l|}{\textbf{CR3} - Contains the physical address of the base of the paging-structure hierarchy and two flags} \\ \hline
  % TODO: Bits 0:11 are assumed to be 0. 3,4 are only used when accessing a PDE in 32-bit paging
  3 & PWT & Page-level Write-Through (not used in PAE or if bit CR4.PCIDE = 1 in long mode) \\
  4 & PCD & Page-level Cache Disable (not used in PAE or if bit CR4.PCIDE = 1 in long mode) \\
  12:31(63) & PDBR & Page Directory Base Register (PAE = Base of PDPT, long = base of PML4T/PML5T) \\
  \hline \multicolumn{3}{|l|}{\textbf{CR4} - Contains flags that enable several architectural extentions} \\ \hline
  0 & VME & Virtual 8086 Mode Extensions \\
  1 & PVI & Protected-mode Virtual Interrupts \\
  2 & TSD & Time Stamp Disable \\
  3 & DE & Debugging Extensions \\
  4 & PSE & Page Size Estension \\
  5 & PAE & Physical Address Extension \\
  6 & MCE & Machine Check Extension \\
  7 & PGE & Page Global Enabled \\
  8 & PCE & Performance-Monitoring Counter enable \\
  9 & OSFXSR & Operating System Support for FXSAVE and FXRSTOR Instructions \\
  10 & OSXMMEXCPT & Operating System Support for Unmasked SIMD Floating-Point Exceptions \\
  11 & UMIP & User-Mode Instruction Prevention (if set, \#GP on SGDT, SIDT, SLDT, SMSW, and STR instructions when CPL $>$ 0) \\
  12 & LA57 & 57-bit linear addresses (if set, the processor uses 5-level paging otherwise it uses 4-level paging) \\
  13 & VMXE & Virtual Machine Extensions Enable \\
  14 & SMXE & Safer Mode Extensions enable \\
  16 & FSGSBASE & Enables the instructions: RDFSBASE, RDGSBASE, WRFSBASE and WRGSBASE \\
  17 & PCIDE & PCID Enable \\
  18 & OSXSAVE & XSAVE and Processor Extended States Enable \\
  20 & SMEP & Supervisor Mode Execution Protection Enable \\
  21 & SMAP & Supervisor Mode Access Prevention Enable \\
  22 & PKE & Protection Key Enable \\
  23 & CET & Control-flow Enforcement Technology \\
  24 & PKS & Enable Protection Keys for Supervisor-Mode Pages \\
  \hline \multicolumn{3}{|l|}{\textbf{CR8} - Provides access to the Task Priority Register (TPR)} \\ \hline
  0:3 & TPL & Task Priority Level \\ \hline
\end{longtable}

\cite[Chapter~2.5]{sdm3}

% TODO: EFLAGS
% https://wiki.osdev.org/CPU_Registers_x86
% https://wiki.osdev.org/CPU_Registers_x86-64

% TODO: Extended control, debug, test, protected mode (gdtr, ldtr, idtr), RFLAGS

\layout{
  \subsubsection{Model Specific Registers}

  Model Specific Registers (MSRs) are registers which may not be present on all processors, some are added for testing features and may not be present in future processors, others may provide unique functionality for the processor (eg if the processor were made to be used in a non-standard environment).
  MSRs are 64-bits in size and identified with a unique 32-bit integer stored in ecx.
  MSRs are witten to using the \texttt{WRMSR} instruction, it writes EAX:EDX to the register;
  MSRs are read from using the \texttt{RDMSR} instruction, it reads from the register into EAX:EDX.

  \bigskip \cite[Chapter~11.4]{sdm3}
} \bigskip

\layout{
  The most commonly used MSR is the Extended Feature Enable Register (EFER, sometimes IA32\_EFER), its address is 0xC000\_0080 and its contents are listed in the table below:
  \begin{table}[H]
    \centering
    \begin{tabular}{|c|c|l|}
      \hline \textbf{Bit} & \textbf{Label} & \textbf{Description} \\ \hline
      0 & SCE & Enables SYSCALL/SYSRET instructions in 64-bit mode \\
      8 & LME & Enabled IA-32e mode \\
      10 & LMA & Indicates IA-32e mode is active when set \\
      11 & NXE & Execute disable bit enable \\ \hline
    \end{tabular}
  \end{table}

  \cite[Page~2-87]{sdm4}
} \bigskip

\layout{
  Various other MSR addresses used in this document (their use is described where they are used):
  \begin{table}[H]
    \centering
    \begin{tabular}{|c|c|l|}
      \hline \textbf{MSR} & \textbf{Address} \\ \hline
      IA32\_MTRR\_DEF\_TYPE & 0x02FF \\
      IA32\_MTRR\_FIX64K\_00000 & 0x0250 \\
      IA32\_MTRR\_FIX16K\_80000 & 0x0258 \\
      IA32\_MTRR\_FIX16K\_A0000 & 0x0259 \\
      IA32\_MTRR\_FIX4K\_C0000 & 0x0268 \\
      IA32\_MTRR\_FIX4K\_C8000 & 0x0269 \\
      IA32\_MTRR\_FIX4K\_D0000 & 0x026A \\
      IA32\_MTRR\_FIX4K\_D8000 & 0x026B \\
      IA32\_MTRR\_FIX4K\_E0000 & 0x026C \\
      IA32\_MTRR\_FIX4K\_E8000 & 0x026D \\
      IA32\_MTRR\_FIX4K\_F0000 & 0x026E \\
      IA32\_MTRR\_FIX4K\_F8000 & 0x026F \\
      IA32\_MTRR\_PHYSBASEn & 0x0202 + 2n \\
      IA32\_MTRR\_PHYSMASKn & 0x0201 + 2n \\
      IA32\_PAT & 0x0227 \\
      \hline
    \end{tabular}
  \end{table}

  \cite{sdm4}
}

\layout{
  \subsection{Cache}

  Cache is a relatively small amount of memory placed within the processor. It is extremely fast SRAM (as opposed to the slower main memory DRAM). There are typically several layers of cache, with each layer getting larger in capacity and slower to access:

  \begin{enumerate}
    \item L1d: \\
      Extremely fast data cache
    \item L1i: \\
      Extremely fast instruction cache
    \item L2: \\
      Very fast cache
    \item L3: \\
      Fast cache (shared between cores)
  \end{enumerate}
}

\layout{
  \subsubsection{Cache coherency protocol}

  Cache coherency protocols ensure the caches on difference processors (cores) store the most recent, correct copy of data.
  x86 uses MSI-based protocols, such as MESI, MOESI and MESIF (with the latter two being used in newer AMD and Intel processors respectively);
  MSI-based protocols are a type of write-invalidate snooping protocol, this means that a write on one cache invalidates all other cache.
  Each cache line is stored with a state, when a cacheline is accessed (read miss or write), its state is updated and it sends a message to all other processors on a bus.
  The cache controllers snoops the bus, if a message involves a cacheline in its cache, its state is updated to reflect the message.
} \bigskip

\layout{
  The accesses to cache which affect state are:
  \begin{enumerate}
    \item PR - current Processor Reads (cache miss)
    \item PW - current Processor Writes
    \item BR - Bus Read (cache miss, another processor reads)
    \item BW - Bus Write (another processor writes)
  \end{enumerate}
}

The state transitions are defined by a finite state machine, the invalid state is considered as a cache miss and uncached items can be considered to be in the invalid state.

\begin{table}[H]
  \centering
  \begin{tabular}{|c|c|c|p{0.5\linewidth}|}
    \hline \textbf{State} & \textbf{Data} & \textbf{Responds to BR} & \textbf{Description} \\ \hline
    Modified & Dirty & Yes & Only copy of the data \\
    Exclusive & Clean & Yes & Only copy of the data \\
    Shared & Clean* & No** & Multiple copies of the data \\
    Invalid & Invalid & No & The cache line does not hold a valid copy of the data \\
    Owned & Dirty & Yes & Multiple copies of the data (all others are shared) \\
    Forward & Clean & Yes & Multiple copies of the data (all others are shared) \\ \hline % https://en.wikipedia.org/wiki/MESIF_protocol
  \end{tabular}
\end{table}

* Dirty if another cache is in the Owned state.

** Will never respond (provide the cached data) if another cache is Owned or Forward, otherwise depending on implementation either a single Shared cache will respond, they all respond or none respond.

Owned and Modified states write to main memory (or next-level cache) when switching to a clean state or when evicted from cache.
Furthermore, states with the only copy of cache (Modified and Exclusive) do not have to send a bus message when writing to its cache.

\layout{
  The following finite state machines represent the transitions between the states of their respective protocols (MSI has been included since it is the foundation of the protocols but is not used in x86).
  \begin{table}[H]
    \centering
    \begin{tabular}{|c|c|c|c|c|}
      \cline{2-5} \multicolumn{1}{c|}{} & \multicolumn{4}{c|}{\textbf{Access}} \\ \hline
      \textbf{State} & \textbf{PR} & \textbf{PW} & \textbf{BR} & \textbf{BW} \\ \hline
      \multicolumn{5}{|c|}{MSI transitions \cite{cache-msi}} \\ \hline
      M & M & M & S & I \\
      S & S & M & S & I \\
      I & S & M & I & I \\ \hline
      \multicolumn{5}{|c|}{MESI transitions \cite{cache-mesi}} \\ \hline
      M & M & M & S & I \\
      E & E & M & S & I \\
      S & S & M & S & I \\
      I & * & M & I & I \\ \hline
      \multicolumn{5}{|c|}{MOESI transitions \cite{cache-moesi}} \\ \hline
      M & M & M & O & I \\
      O & O & M & O & I \\
      E & E & M & S & I \\
      S & S & M & S & I \\
      I & * & M & I & I \\ \hline
      \multicolumn{5}{|c|}{MESIF transitions \cite{cache-mesif}} \\ \hline
      M & M & M & S & I \\
      E & E & M & S & I \\
      S & S & M & S & I \\
      I & ** & M & I & I \\
      F & F & M & S & I \\ \hline
    \end{tabular}
  \end{table}

  * Transitions to E if exclusive or S if shared.

  ** Transitions to E if exclusive or F if shared.
}

\layout{
  \subsubsection{Cache Placement Policy}

  Cache placement policies determine which cacheline to place a block of memory, it is required for efficient lookup and utilisation of cache.
} \bigskip

\layout{
  There are three main cache placement policies:
  \begin{itemize}
    \item Direct: \\
      Each physical address is split into three parts: tag, block and offset.
      The block is used to identify a specific cacheline, each cacheline stores a tag and the tags are compared, if equal its considered a cache hit else it is a cache miss.
      On a hit, the offset is used to retrieve a byte from the cacheline.
      On a miss, the new data is fetched from main memory and replaces the cacheline.
    \item Associative: \\
      Each physical address is split into two parts: tag and offset.
      Every cacheline is checked in parallel for if the tags are equal.
      On a hit, the offset is used to retrieve a byte from the cacheline.
      On a miss, the new data is fetched from main memory and replaces a cacheline usually identified by a LRU algorithm (Least Recently Used).
    \item N-Way Set-Associative: \\
      Each physical address is split into three parts: tag, block and offset.
      The block is used to identify a specific set (each set stores N cachelines).
      Each cacheline in the set is checked in parallel for if the tags are equal.
      On a hit, the offset is used to retrieve a byte from the cacheline.
      On a miss, the new data is fetched from main memory and replaces a cacheline within the set usually identified by a LRU algorithm.
  \end{itemize}
}

\layout{
  The Following table summarises the pros and cons of each policy:
  \begin{table}[H]
    \centering
    \begin{tabular}{|>{\centering\arraybackslash}m{0.15\linewidth}|m{0.25\linewidth}|m{0.25\linewidth}|}
      \hline \textbf{Policy} & \textbf{Pros} & \textbf{Cons} \\ \hline
      Direct & Fast (spacial locality) & Cache lines often get evicted if they have the same block \\ \hline
      Associative & Cache lines only get evicted if not used (temporal locality) & Slow/Expensive \\ \hline
      Set-Associative & Faster than Associative with similar positives & Slower than Direct and Conflict misses can occur \\ \hline
    \end{tabular}
  \end{table}

  \cite{cache-placement}

  \bigskip

  Most x86 processors use N-Way Set-Associative caches, with the more performant processors using larger set sizes.
}

\layout{
  \subsubsection{Cache misses}

  A cache miss occurs when valid data does not exist in the processor's cache, requiring a read from main memory.
  The first three types of cache miss are commonly known as the three C's.

  \begin{enumerate}
    \item Compulsory (cold) miss: \\
      Occurs when accessing memory for the first time (so was never cached)
    \item Capacity miss: \\
      Occurs when cache is full and a cacheline was evicted but later used
    \item Conflict miss: \\
      Occurs when multiple cache lines are mapped to the same location in cache, causing one to get evicted so if read again later it is a miss
    \item Coherence miss: \\
      The (write-invalidate) cache coherency protocol (eg MESI) invalidated the cacheline because it was written in another processor
  \end{enumerate}

  \cite{cache-miss}
}

\layout{
  \subsubsection{Memory Types}

  Memory types define the caching and write back rules for a region of memory.
}

\begin{table}[H]
  \centering
  \begin{tabular}{|l|c|>{\centering}m{0.12\linewidth}|>{\centering}m{0.13\linewidth}|c|}
    \hline \textbf{Memory Type} & \textbf{Cacheable} & \textbf{Write-Back Cacheable} & \textbf{Allows Speculative Reads} & \textbf{Memory Ordering Model} \\ \hline
    Uncacheable (UC) & No & No & No & Strong Ordering \\
    Write Combining (WC) & No & No & Yes & Weak Ordering \\
    Write Through (WT) & Yes & No & Yes & Speculative Processor Ordering \\
    Write Back (WB) & Yes & Yes & Yes & Speculative Processor Ordering \\
    Write Protected (WP) & Reads & No & Yes & Speculative Processor Ordering \\ \hline
  \end{tabular}
\end{table}

\layout{
  Additional notes for the memory types:
  \begin{itemize}
    \item UC: Useful for memory-mapped IO, worst performance.
    \item WC: Writes may be delayed and combined in the write combining buffer (WC buffer) to reduce memory accesses.
    \item WT: Writes go to main memory (updating any existing cache lines), write combining is allowed.
    \item WB: Writes go to cache, write combining is allowed, best performance.
    \item WP: Writes go to main memory and invalidate all cache lines.
  \end{itemize}

  \cite[Chapter~13.3]{sdm3}
}

\layout{
  \subsubsubsection{MTRRs}

  Memory Type Range Registers (MTRRs) are used to define ranges in physical memory, various MSRs specify the memory type in each range.
} \bigskip

\layout{
  Memory types that can be encoded with MTRRs:
  \begin{table}[H]
    \centering
    \begin{tabular}{|c|c|c|c|c|}
      \hline \textbf{Encoding} & \textbf{Memory Type} \\ \hline
      00 & UC \\
      01 & WC \\
      02 & Rsvd. \\
      03 & Rsvd. \\
      04 & WT \\
      05 & WP \\
      06 & WB \\
      07-FF & Rsvd. \\ \hline
    \end{tabular}
  \end{table}
}

\layout{
  The IA32\_MTRR\_DEF\_TYPE MSR sets the default memory type for memory regions not specified by an MTRR and is also used to enable MTRRs.
  \begin{table}[H]
    \centering
    \begin{tabular}{|c|c|}
      \hline \textbf{Bits} & \textbf{Use} \\ \hline
      7:0 & Default memory type \\
      10 & Fixed-range MTRRs enable \\
      11 & (all) MTRRs enable \\ \hline
    \end{tabular}
  \end{table}
}

\layout{
  Fixed Range MTRRs are a type of MTRR which operate on a fixed region of memory.
  There are 11 fixed-range registers of 64 bytes each, each is divided into eight 8-bit fields used to specify the memory type for the sub-ranges of the register.
  The IA32\_MTRR\_FIX64K\_00000 MSR divides the 512-KiB range 7FFFF:00000 into eight 64-KiB sub-ranges.
  The IA32\_MTRR\_FIX16K\_80000 and IA32\_MTRR\_FIX16K\_A0000 MSRs divide the two 128-KiB ranges within the range BFFFF:80000 into sixteen 16-KiB sub-ranges (8 per register).
  The IA32\_MTRR\_FIX4K\_C0000 through IA32\_MTRR\_FIX4K\_F8000 MSRs divide the eight 32-KiB ranges within the range FFFFF:C0000 into sixty-four 4-KiB sub-ranges (8 per register).
} \bigskip

\layout{
  Variable Range MTRRs are a type of MTRR which operate on a variable region of memory.
  The number (m) of variable range MTRRs supported is given in bits 7:0 of the IA32\_MTRRCAP MSR.
  The entry in IA32\_MTRR\_PHYSBASEn defines the base address and memory type for the range,
  the entry in IA32\_MTRR\_PHYSMASKn contains a mask used to determine the range,
  where n is an integer in the range 0 to m-1.
} \bigskip

\layout{
  Layout of IA32\_MTRR\_PHYSBASEn MSR:
  \begin{table}[H]
    \centering
    \begin{tabular}{|c|c|}
      \hline \textbf{Bits} & \textbf{Use} \\ \hline
      7:0 & Memory type \\
      12:MAXPHYADDR-1 & Base address \\ \hline
    \end{tabular}
  \end{table}
}

\layout{
  Layout of IA32\_MTRR\_PHYSMASKn MSR:
  \begin{table}[H]
    \centering
    \begin{tabular}{|c|c|}
      \hline \textbf{Bits} & \textbf{Use} \\ \hline
      11 & Valid (enables the MTRR entry) \\
      12:MAXPHYADDR-1 & Mask \\ \hline
    \end{tabular}
  \end{table}
}

\layout{
  An address is within the range of a variable MTRR when: \texttt{Address AND Mask = Base AND Mask}.
  Furthermore there are STRRs (System-Management Range Registers) used in System Management Mode.

  \bigskip \cite[Chapter~13.11.2]{sdm3}
} \bigskip

\layout{
  The MTRR memory type is determined as follows:
  \begin{itemize}
    \item If MTRRs are not enabled then all memory accesses are UC.
    \item Else if the physical address is within the first MiB and fixed MTRRs are enabled then it uses the appropriate fixed-range MTRR.
    \item Else if the address falls within a variable MTRR then it is used.
      If it falls within multiple variable MTRRs and they are not identical then, UC is used if one is UC, WT is used if they are all WT or WB otherwise it is undefined.
    \item Otherwise it uses the default.
  \end{itemize}
  \cite[Chapter~13.11.4.1]{sdm3}
}

\layout{
  \subsubsubsection{Page Attribute Table}

  The Page Attribute Table (PAT) provides additional functionality to set the memory type for page entries.

  \cite[Chapter~13.12]{sdm3} \bigskip

  Furthermore, PAT is supported on all 64-bit processors.

  \cite[Chapter~5.9.1]{sdm3}
} \bigskip

\layout{
  Memory Types that can be encoded with PAT:
  \begin{table}[H]
    \centering
    \begin{tabular}{|c|c|}
      \hline \textbf{Encoding} & \textbf{Memory Type} \\ \hline
      00 & UC \\
      01 & WC \\
      02 & Rsvd. \\
      03 & Rsvd. \\
      04 & WT \\
      05 & WP \\
      06 & WB \\
      07 & UC- \\
      08-FF & Rsvd. \\ \hline
    \end{tabular}
  \end{table}
}

\layout{
  The IA32\_PAT MSR stores eight PAT entries, each of which stores a memory type and can be selected by page entries.

  \bigskip \cite[Chapter~13.12.2]{sdm3}
} \bigskip

\layout{
  PAT Entries:
  \begin{table}[H]
    \centering
    \begin{tabular}{|c|c|c|c|c|}
      \hline \textbf{PAT} & \textbf{PCD} & \textbf{PWT} & \textbf{PAT Entry} & \textbf{Default Memory Type} \\ \hline
      0 & 0 & 0 & PAT0 & WB \\
      0 & 0 & 1 & PAT1 & WT \\
      0 & 1 & 0 & PAT2 & UC- \\
      0 & 1 & 1 & PAT3 & UC \\
      1 & 0 & 0 & PAT4 & WB \\
      1 & 0 & 1 & PAT5 & WT \\
      1 & 1 & 0 & PAT6 & UC- \\
      1 & 1 & 1 & PAT7 & UC \\ \hline
    \end{tabular}
  \end{table}
}

\layout{
  Updating the PAT usually requires the OS to invalidate all affected TLB entries

  \bigskip \cite[Chapter~13.12.3]{sdm3} \cite[Chapter~13.12.4]{sdm3}
}

\layout{
  \subsubsubsection{Effective Memory Type}

  The effective memory type is based on a combination of the MTRR memory type and the PAT (or PCD and PWT if PAT is unsupported) for the memory address.
  If Paging is not used, then the effective memory type is that of the MTRR.
  Furthermore if CR0.CD is set then caching is prevented.

  \bigskip \cite[Chapter~13.5.2]{sdm3}
} \bigskip

\layout{
  Effective Memory Type (PAT Unsupported):
  \begin{table}[H]
    \centering
    \begin{tabular}{|c|c|c|c|c|c|c|}
      \hline \multicolumn{2}{|c|}{\textbf{Page Flags}} & \multicolumn{5}{c|}{\textbf{MTRR}} \\ \hline
      \textbf{PCD} & \textbf{PWT} & \textbf{UC} & \textbf{WC} & \textbf{WT} & \textbf{WB} & \textbf{WP} \\ \hline
      \textbf{0} & \textbf{0} & UC & WC & WT & WP & WB \\ \hline
      \textbf{0} & \textbf{1} & UC & WC* & WT & WP* & WT \\ \hline
      \textbf{1} & \textbf{0} & UC & WC* & UC & WC* & UC \\ \hline
      \textbf{1} & \textbf{1} & UC & UC & UC & UC & UC \\ \hline
    \end{tabular}
  \end{table}

  * Implementation defined, avoid these combinations.

  \bigskip \cite[Chapter~13.5.2.1]{sdm3}
} \bigskip

\layout{
  Effective Memory Type (PAT Supported):
  \begin{table}[H]
    \centering
    \begin{tabular}{c|c|c|c|c|c|c|}
      \multicolumn{2}{c}{} & \multicolumn{5}{c}{\textbf{MTRR}} \\ \cline{3-7}
      \multicolumn{2}{c|}{} & \textbf{UC} & \textbf{WC} & \textbf{WT} & \textbf{WB} & \textbf{WP} \\ \cline{2-7}
      \multirow[c]{6}*{\textbf{PAT}} & \textbf{UC} & UC1 & UC2 & UC2 & UC2 & UC2 \\ \cline{2-7}
      & \textbf{UC-} & UC1 & WC & UC2 & UC2 & WC \\ \cline{2-7}
      & \textbf{WC} & WC & WC & WC & WC & WC \\ \cline{2-7}
      & \textbf{WT} & UC1 & UC2 & WT & WT & WT \\ \cline{2-7}
      & \textbf{WB} & UC1 & WC & WT & WB & WP \\ \cline{2-7}
      & \textbf{WP} & UC1 & UC2 & WP & WP & WP \\ \cline{2-7}
    \end{tabular}
  \end{table}

  UC1 - UC attribute comes from MTRRs so processors are not required to snoop their caches since the data could never have been cached. Preferred over UC2 for performance Reasons.

  UC2 - UC attribute comes from PAT so processors are required to snoop their caches since the data may be cached due to page aliasing (which is not recommended).

  \bigskip \cite[Chapter~13.5.2.2]{sdm3}
}

\layout{
  \subsection{Memory addressing}

  Memory addresses are used to reference specific locations in memory (which can be accessed by dereferencing the address), (effective) memory addresses in assembly code have the syntax: \texttt{[Base + Index * Scale + Displacement]}.
  Furthermore a segment override can be added using the syntax: \texttt{Segment:[...]}.
} \bigskip

\layout{
  x86 has different types of memory address:
  \begin{enumerate}
    \item Logical: \\
      Consists of a segment selector (16-bit) and an offset (same bit size as the code).
      Usually inferred from the type of memory access (eg uses CS for code, SS for stack) or from a segment override.
      Produces a linear address after segmentation.
    \item Effective: \\
      Consists of a combination of a \texttt{Base} register, an \texttt{Index} register (optionally with an immediate \texttt{Scale} which is 1, 2, 4, or 8) and an immediate \texttt{Displacement}.
      Once the effective address is computed it is used as the offset in the logical address.
    \item Linear: \\
      If paging is not used, the linear address is equal to the physical address.
      If paging is used, the corresponding physical address is found using the paging hierarchy.
    \item Virtual: \\
      A valid linear address that does not correspond to an in-memory page, causing the kernel to generate the page via an interrupt (eg load it from disk).
    \item Physical: \\
      The physical location of the memory in main memory.
  \end{enumerate}
}

% TODO: not here, idk where: priviledge levels: checking for instructions, managing (kernel)
% https://stackoverflow.com/questions/20155304/does-linux-use-x86-cpus-pcid-feature-for-tlb-if-not-why

\layout{
  \subsection{Segmentation}

  Segmentation is a method of memory management that separates parts of a processes address space into different segments (such as code and data), it uses the 6 segment registers to keep track of the segments.
  It is used to convert a logical address (which consists of a segment register and an offset) into a linear address.
  Segmentation has different behaviours depending on what operating mode the processor is in, unlike paging it is always enabled.

  \bigskip \cite[Chapter~3.1]{sdm3}
}

\layout{
  \subsubsection{Real Mode}

  The value in the segment register is left shifted by 4 (multiplied by 16) then added to the offset to produce a 20-bit physical address allowing up to 1MiB of memory instead of the 16-bit limit of 64KiB \cite[Chapter~22.1.1]{sdm3}.
  The kernel allocates memory for each running process and gives them each different segments, this allows the programs to be able to be loaded into any part of the address space and to not affect other program's memory.
  However there is no hardware enforcement of memory protection as any process can read and modify any memory location.
}

\layout{
  \subsubsection{Protected Mode}

  The value in the segment register is a segment selector, it is used as an index into the LDT/GDT to select a segment descriptor.
  Each segment descriptor contains a base address, limit and flags.
  The base address is a linear address and points to the base of the segment, the limit is the size of the segment and the flags set the segments properties such as permissions, 32 or 16 bit code and granularity of the limit.
  There are many models which can be used to implement segmentation; some common models are listed below, other models exist and additional functionality can be added to the models such using the code segment to point to ROM.

  \bigskip \cite[Chapter~3.1]{sdm3}
}

\begin{enumerate}
  \item Basic Flat Model \cite[Chapter~3.2.1]{sdm3}: \\
    Contains at least two segment descriptors, one for code and one for data (this is because the code descriptor has to be executable so is usually read-only).
    Both descriptors are mapped to the entire linear address space (even if the physical memory does not exist) so it has a base address of 0 and a limit of 4GiB.
    This essentially disables segmentation for both the application and system developer, and since the entire linear address space is mapped no \#GP exceptions will occur.
    Usually implemented alongside paging.
  \item Protected Flat Model \cite[Chapter~3.2.2]{sdm3}: \\
    Similar to the Basic Flat Model but only included the range of addresses for which physical memory actually exists, so the limit is set to the capacity of physical memory.
    If a process accesses a non-existent memory location a \#GP exception will occur.
  \item Multi-Segment Model \cite[Chapter~3.2.3]{sdm3}: \\
    Uses the full capabilities of segmentation to provide hardware enforced protection of memory.
    Each program is given its own segment descriptor table and multiple segments (which may be private or shared with other programs).
    Hardware checks can be used to enforce the program to access memory it owns and in a way the OS allows (such as not writing to code/read-only memory)
\end{enumerate}

\layout{
  \subsubsection{Long Mode}

  The behaviour of segmentation in long mode depends if it is running in 64-bit or compatibility mode.
  In compatibility mode it functions the same as it does in protected mode.
  In 64-bit mode, segmentation is generally disabled (but not completely), which creates a flat 64-bit linear address space.
  The base address of the CS, DS, ES and SS are treated as 0 (the effective address is equal to the linear address), and the processor does not perform limit checks (on all segment registers).
  The FS and GS registers are often used to reference thread-local or kernel data structures since their base address still applies.

  \bigskip \cite[Chapter~3.2.4]{sdm3}
}

\layout{
  \subsubsection{Segment Selectors}

  Segment selectors are used as a 16-bit identifier for a segment descriptor.
} \bigskip

\layout{
  Segment selectors contain 3 parts:
  \begin{enumerate}
    \item Bits 15:3 - Index: \\
      The index into the descriptor table, selects one of the 8192 segment descriptors in the table.
      Its value is multiplied by 8 then added to the base address of the of the descriptor table.
    \item Bit 2 - Table Indicator (TI) flag: \\
      Selects between the two descriptor tables: 0 selects the GDT to be used, 1 selects the LDT to be used.
    \item Bits 1:0 - Requested Privilege Level (RPL): \\
      Specifies the privilege level of the selector.
  \end{enumerate}
}

\layout{
  If the index is 0, it implicitly selects a null segment.
  Additionally both the CS and SS registers have to be loaded with a present descriptor else a \#GP exception is generated; furthermore if the SS is loaded with a non-writeable segment, a \#GP is generated.

  \bigskip \cite[Chapter~3.4.2]{sdm3}
}

\layout{
  \subsubsection{Segment Descriptors}

  Segment descriptors are identifiers for segments of memory.
  They contain multiple parts which can be categorised into: (linear) base address of the segment, limit (size of segment - 1), and flags (controls the abilities of the segment).
} \bigskip

\layout{
  Layout of segment descriptors:
  \begin{table}[H]
    \centering
    \begin{tabular}{|c|c|l|l|}
      \hline \textbf{Bits} & \textbf{Size} & \textbf{Name} & \textbf{Use} \\ \hline
      63:56 & 8 & Base 31:24 & 1 byte of the base address \\
      55:55 & 1 & Granularity (G) & 0 = limit in bytes, 1 = limit in 4-KiB \\
      54:54 & 1 & D/B & Default operation size, 0 = 16-bit, 1 = 32-bit \\
      53:53 & 1 & L & 1 = 64-bit code segment (IA-32e mode only) \\
      52:52 & 1 & AVL & Available for use by system software \\
      51:48 & 4 & Limit 19:16 & 4 bits of the limit \\
      47:47 & 1 & Present (P) & 0 = segment not present, 1 = segment present \\
      46:45 & 2 & DPL & Descriptor privilege level \\
      44:44 & 1 & S & Descriptor type, 0 = system, 1 = code/data \\
      43:40 & 4 & Type & Segment type \\
      39:16 & 24 & Base 23:00 & 3 byte of the base address \\
      15:00 & 16 & Limit 15:00 & 2 bytes of the base address \\ \hline
    \end{tabular}
  \end{table}

  \cite[Chapter~3.4.5]{sdm3}
} \bigskip

\layout{
  Segment types:
  \begin{table}[H]
    \centering
    \begin{tabular}{|c|c|c|c|c|p{0.303\linewidth}|p{0.303\linewidth}|}
      \cline{2-5} \multicolumn{1}{c|}{} & \multicolumn{4}{c|}{\textbf{Type}} & \multicolumn{2}{c}{} \\ \hline
      \textbf{S} & \textbf{3} & \textbf{2} & \textbf{1} & \textbf{0} & \multicolumn{1}{c|}{\textbf{32-bit mode}} & \multicolumn{1}{c|}{\textbf{IA-32e mode}} \\ \hline
      0 & 0 & 0 & 0 & 0 & Reserved & Reserved \\
      0 & 0 & 0 & 0 & 1 & 16-bit TSS (Available) & Reserved \\
      0 & 0 & 0 & 1 & 0 & LDT & LDT \\
      0 & 0 & 0 & 1 & 1 & 16-bit TSS (Busy) & Reserved \\
      0 & 0 & 1 & 0 & 0 & 16-bit Call Gate & Reserved \\
      0 & 0 & 1 & 0 & 1 & Task Gate & Reserved \\
      0 & 0 & 1 & 1 & 0 & 16-bit Interrupt Gate & Reserved \\
      0 & 0 & 1 & 1 & 1 & 16-bit Trap Gate & Reserved \\
      0 & 1 & 0 & 0 & 0 & Reserved & Reserved \\
      0 & 1 & 0 & 0 & 1 & 32-bit TSS (Available) & 64-bit TSS (Available) \\
      0 & 1 & 0 & 1 & 0 & Reserved & Reserved \\
      0 & 1 & 0 & 1 & 1 & 32-bit TSS (Busy) & 64-bit TSS (Busy) \\
      0 & 1 & 1 & 0 & 0 & 32-bit Call Gate & 64-bit Call Gate \\
      0 & 1 & 1 & 0 & 1 & Reserved & Reserved \\
      0 & 1 & 1 & 1 & 0 & 32-bit Interrupt Gate & 64-bit Interrupt Gate \\
      0 & 1 & 1 & 1 & 1 & 32-bit Trap Gate & 64-bit Trap Gate \\ \hline
      \textbf{-} & \textbf{-} & \textbf{E} & \textbf{W} & \textbf{A} & \multicolumn{1}{c|}{\textbf{Type}} & \multicolumn{1}{c|}{\textbf{Description}} \\ \hline
      1 & 0 & 0 & 0 & 0 & Data & Read-Only \\
      1 & 0 & 0 & 0 & 1 & Data & Read-Only, accessed \\
      1 & 0 & 0 & 1 & 0 & Data & Read-Write \\
      1 & 0 & 0 & 1 & 1 & Data & Read-Write, accessed \\
      1 & 0 & 1 & 0 & 0 & Data & Read-Only, expand-down \\
      1 & 0 & 1 & 0 & 1 & Data & Read-Only, expand-down, accessed \\
      1 & 0 & 1 & 1 & 0 & Data & Read-Write, expand-down \\
      1 & 0 & 1 & 1 & 1 & Data & Read-Write, expand-down, accessed \\ \hline
      \textbf{-} & \textbf{-} & \textbf{C} & \textbf{R} & \textbf{A} & \multicolumn{1}{c|}{\textbf{Type}} & \multicolumn{1}{c|}{\textbf{Description}} \\ \hline
      1 & 1 & 0 & 0 & 0 & Code & Execute-Only \\
      1 & 1 & 0 & 0 & 1 & Code & Execute-Only, accessed \\
      1 & 1 & 0 & 1 & 0 & Code & Execute-Read \\
      1 & 1 & 0 & 1 & 1 & Code & Execute-Read, accessed \\
      1 & 1 & 1 & 0 & 0 & Code & Execute-Only, conforming \\
      1 & 1 & 1 & 0 & 1 & Code & Execute-Only, conforming, accessed \\
      1 & 1 & 1 & 1 & 0 & Code & Execute-Read, conforming \\
      1 & 1 & 1 & 1 & 1 & Code & Execute-Read, conforming, accessed \\ \hline
    \end{tabular}
  \end{table}
}

\layout{
  Description of bits used by the segment types:
  \begin{enumerate}
    \item E = Expansion-direction: \\
      If 0, the segment is expand-up;
      this means that the range of valid offsets is [0, limit].
      Therefore the range of linear addresses of the segment is [base, base + limit], increasing the limit increases the maximum address (useful for data segments).
      If 1, the segment is expand-down;
      this means that the range of valid offsets is (limit, 0xFFFFF] if B = 0 or (limit, 0xFFFFFFFF] if B = 1.
      Therefore the range of linear addresses of the segment is (base + limit, base + (0xFFFFF or 0xFFFFFFFF)], decreases the limit decreases the minimum address (useful for stack segments).
    \item W = Write-enabled: \\
      Allows the segment to be written; the stack segment must be writeable, loading the SS register with a read only segment causes a \#GP exception.
    \item A = Accessed: \\
      The processor sets it to 1 if it accessed the segment.
      The kernel can clear its value and it is often useful for debugging.
      If the segment descriptor is placed in ROM, the processor may enter an infinite loop attempting to set the Accessed bit.
      To prevent this, the segment descriptors in ROM should have their Accessed bit set.
    \item C = Conforming: \\
      A transfer of execution into a more privileged conforming segment will continue at the current privilege level.
      Transferring execution into a more privileged non-conforming segment or less privileged segment will result in a \#GP exception if a call or task gate is not used. % TODO: Check that a high to low privilege will prevent a GP if a gate is used
      All data segments are non-conforming so cannot be accessed by less privileged processes, however unlike code segments higher privileged processes can access lower privileged data segments without the use of a gate.
    \item R = Read-enabled: \\
      Allows the code segment to be read and is useful if constants or static data are placed within the code segment.
      The code segments can be read either by using a CS override or by loading the segment into a data-segment register (DS, ES, GS, FS).
  \end{enumerate}

  \cite[Chapter~3.4.5]{sdm3} \cite[Chapter~3.4.5.1]{sdm3} \cite[Chapter~3.5]{sdm3}
} \bigskip

Furthermore, when a segment register is loaded with a segment selector, the processor stores the corresponding descriptor in the hidden part of the segment register (the descriptor cache), if the descriptor is changed it is the responsibility of the software to reload the segment selector \cite[Chapter~3.4.3]{sdm3}.

\bigskip

Additionally in IA-32e mode, system descriptors are expended to 16 bytes (occupying the space of two entries) \cite[Chapter~3.52]{sdm3}.

\layout{
  \subsubsection{Segment Descriptor Tables}

  The Global Descriptor Table (GDT) contains segment descriptors, its base address (linear) and limit (number of bytes in the GDT - 1) are stored in the GDTR (GDT Register).
  The \texttt{LGDT} instruction stores the base address and limit from a memory location into the GDTR.
  The $0^{th}$ descriptor of the GDT is often used to store the address and limit since it will never be loaded as a descriptor since the selector would be null.
  There is usually only a single GDT used by all processes.
} \bigskip

\layout{
  The Local Descriptor Table (LDT) also contains segment descriptors; however unlike the GDT there are usually many LDTs (in a non-flat segmentation model), this is so that each process can have its own set of segments.
  Unlike the GDTR, the LDTR (LDT Register) stores the segment selector for the LDT table segment stored in the GDT.
  The \texttt{LLDT} instruction stores the segment selector of the new LDT in the LDTR.
  Furthermore in 32-bit task switches the LDTR is automatically loaded with the segment selector and descriptor for the LDT of the new task.

  \bigskip \cite[Chapter~3.5.1]{sdm3}
}

% TODO: check if these are needed: sections, tss

\layout{
  \subsection{Paging}

  % TODO: Cite this first bit here

  Paging is a method of memory management that separates the linear address space from the physical address space.
  It occurs after segmentation but unlike segmentation it does not need to be enabled, if it is disabled linear addresses are the same as physical addresses.
  The OS splits up the physical address space into pages, each process has a paging hierarchy which the OS manages, it consists of many paging structures.
  When a process accesses memory, the processor translates the linear address into its corresponding physical address using the paging hierarchy (by traversing the paging hierarchy), this is called a hardware-assisted page walk.
  There are four different paging modes.
}

\layout{
  \subsubsection{Enabling Paging}

  Before paging is enabled, CR3 should contain the physical address of the first paging structure, and protection (CR0.PE) is enabled.
  Furthermore, once paging is enabled a different paging mode can be used but paging should be disabled first before changing flags since exceptions may occur by changing the flags whilst paging is enabled.
  Paging is enabled by setting CR0.PG.

  \bigskip \cite[Chapter~5.1.2]{sdm3}
} \bigskip

\layout{
  The following table shows which combination of flags is used to enable which paging mode:
  \begin{table}[H]
    \centering
    \begin{tabular}{|c|c|c|c|}
      \hline \textbf{Mode} & \textbf{CR4.PAE} & \textbf{IA32\_EFER.LME} & \textbf{CR4.LA57} \\ \hline
      32-bit & 0 & 0 & -  \\
      PAE & 1 & 0 & -  \\
      4-level & 1 & 1 & 0 \\
      5-level & 1 & 1 & 1 \\ \hline
    \end{tabular}
  \end{table}
}

\layout{
  Since IA32\_EFER.LME enables long mode (IA-32e), 32-bit and PAE can only be used in legacy protected mode and 4-level and 5-level can only be used in long mode.
  Furthermore, in compatibility mode (sub mode of IA-32e), bits 63-32 of the linear address are treated as 0.

  \bigskip \cite[Chapter~5.1.1]{sdm3}
}

\layout{
  \subsubsection{Paging Structures}

  All four paging modes translate linear addresses using hierarchical paging structures.
  Each structure is 4096 bytes (4-KiB) in size and comprises several entries,
  in 32-bit paging each entry is 32 bits (4 bytes) hence there are 1024 entries in each structure,
  in the other paging modes each entry is 64 bits (8 bytes) hence there are 512 entries in each structure.
  However in PAE paging, there is a structure 32 bytes in size containing 4 64-bit entries.
} \bigskip

\layout{
  The processor uses the upper portion of a linear address to identify a series of paging-structure entries. The last of these entries identifies the physical address of the region to which the linear address translates (called the page frame). The lower portion of the linear address (called the page offset) identifies the specific address within that region to which the linear address translates.
  Each paging structure entry contains a physical address which is either the address of another paging structure or a page frame. In the first case it is said to reference the other paging structure; in the second the entry is said to map a page.
  CR3 contains the physical address of the first paging structure (unless HLAT paging is used).
} \bigskip

\layout{
  Linear addresses are translated into physical addresses using an iterative procedure:
  a portion of the linear address (initially the uppermost bits) is used to select an entry in a paging structure (initially the one referenced in CR3), if the entry references another structure then the process continues using another portion of the linear address (immediately below the portion just used), if the entry maps a page then the remaining lower portion of the linear address is used as a page offset.
} \bigskip

\layout{
  A paging-structure entry is marked as not present if its P flag (bit 0) is clear, if it is encountered during translation it causes a page-fault exception; if the exception handler returns, the instruction that generated the exception is restarted.
  Usually an address is mapped to a 4 KiB page, if it is not the final paging structure then bit 7 (PS - Page Size) is checked;
  if PS = 0 then it references a page structure, else if PS = 1 then the entry maps a larger page (4MiB for 32-bit paging if CR4.PSE = 1, 2MiB for other paging modes regardless of CR4.PSE; 4-level and 5-level paging may also support 1GiB pages),
} \bigskip

\cite[Chapter~5.2]{sdm3} \bigskip

\layout{
  The following table lists paging structures along with: the structure that stores their physical address (CR3/previous structure), the bits of the linear address that is used to select the entry within them, and the size of the page they map to (if they map to a page).

  \newcommand{\pgtable}[2]{\multicolumn{1}{>{\raggedright\arraybackslash}m{#1\linewidth}|}{\textbf{#2}}}

  \begin{table}[H]
    \centering
    \begin{tabular}{|p{0.165\linewidth}|p{0.09\linewidth}|l|p{0.105\linewidth}|p{0.14\linewidth}|p{0.18\linewidth}|}
      % \hline \textbf{Paging Structure} & \textbf{Entry Name} & \textbf{Paging Mode} & \textbf{Physical Address of Structure} & \textbf{Bits Selecting Entry} & \textbf{Page Mapping} \\ \hline
      \hline \textbf{Paging Structure} & \pgtable{0.09}{Entry Name} & \textbf{Paging Mode} & \pgtable{0.105}{Physical Address of Structure} & \pgtable{0.14}{Bits Selecting Entry} & \textbf{Page Mapping} \\ \hline
      \multirow[l]{2}{\linewidth}{PML5 table} & \multirow[l]{2}{\linewidth}{PML5E} & 32-bit, PAE, 4-level & \multicolumn{3}{c|}{N/A} \\ \cline{3-6}
       & & 5-level & CR3 & 56:48 & N/A (PS must be 0) \\ \hline
      \multirow[l]{3}{\linewidth}{PML4 table} & \multirow[l]{3}{\linewidth}{PML4E} & 32-bit, PAE & \multicolumn{3}{c|}{N/A} \\ \cline{3-6}
        & & 4-level & CR3 & \multirow[l]{2}{\linewidth}{47:39} & \multirow[l]{2}{\linewidth}{N/A (PS must be 0)} \\ \cline{3-4}
        & & 5-level & PML5E & & \\ \hline
      \multirow[l]{3}{\linewidth}{Page-directory-pointer table} & \multirow[l]{3}{\linewidth}{PDPTE} & 32-bit & \multicolumn{3}{c|}{N/A} \\ \cline{3-6}
        & & PAE & CR3 & 31:30 & N/A (PS must be 0) \\ \cline{3-6}
        & & 4-level, 5-level & PML4E & 38:30 & 1-GiB page if PS=1 \\ \hline
      \multirow[l]{2}{\linewidth}{Page directory} & \multirow[l]{2}{\linewidth}{PDE} & 32-bit & CR3 & 31:22 & 4-MiB page if PS=1 \\ \cline{3-6}
       & & PAE, 4-level, 5-level & PDPTE & 29:21 & 2-MiB page if PS=1 \\ \hline
      \multirow[l]{2}{\linewidth}{Page table} & \multirow[l]{2}{\linewidth}{PTE} & 32-bit & \multirow[l]{2}{\linewidth}{PDE} & 21:12 & \multirow[l]{2}{\linewidth}{4-KiB page} \\ \cline{3-3} \cline{5-5}
       & & PAE, 4-level, 5-level & & 20:12 & \\ \hline
    \end{tabular}
  \end{table}

  \cite[Chapter~5.2]{sdm3}
} \bigskip

The following tables describe the layout of the paging structures (M = MAXPHYADDR, * = Pointer to next paging structure or page frame).

\bigskip

\setlength{\tabcolsep}{0pt}

\layout{
  32-bit Paging:
  \begin{table}[H]
    \centering
    \begin{tabular}{|*{6}{>{\centering\arraybackslash}m{0.13\linewidth}|}}
      \hline \textbf{Bits} & \textbf{CR3} & \textbf{PDE: 4MB} & \textbf{PDE: PTE} & \textbf{PTE: 4KB} & \textbf{Not Present} \\ \hline
      31:22 & \multirow{4}*{PD*} & 31:22 Page* & \multirow{4}*{PT*} & \multirow{4}*{Page*} & \multirow{15}*{Ign.} \\ \cline{1-1} \cline{3-3}
      21:17 & & Rsvd. & & & \\ \cline{1-1} \cline{3-3}
      16:13 & & 39:32 Page* & & & \\ \cline{1-1} \cline{3-3}
      12 & & PAT & & & \\ \cline{1-5}

      11 & \multirow{7}*{Ign.} & \multirow{3}*{Ign.} & \multirow{4}*{Ign.} & \multirow{3}*{Ign.} & \\ \cline{1-1}
      10 & & & & & \\ \cline{1-1}
      9 & & & & & \\ \cline{1-1} \cline{3-3} \cline{5-5}
      8 & & G & & G & \\ \cline{1-1} \cline{3-5}
      7 & & 1 & 0 & PAT & \\ \cline{1-1} \cline{3-5}
      6 & & D & Ign. & D & \\ \cline{1-1} \cline{3-5}
      5 & & A & A & A & \\ \cline{1-5}
      4 & PCD & PCD & PCD & PCD & \\ \cline{1-5}
      3 & PWT & PWT & PWT & PWT & \\ \cline{1-5}
      2 & \multirow{3}*{Ign.} & U/S & U/S & U/S & \\ \cline{1-1} \cline{3-5}
      1 & & R/W & R/W & R/W & \\ \cline{1-1} \cline{3-6}
      0 & & 1 & 1 & 1 & 0 \\ \hline
    \end{tabular}
  \end{table}

  \cite[Chapter~5.3]{sdm3}
} \bigskip

\layout{
  PAE Paging:
  \begin{table}[H]
    \centering
    \begin{tabular}{|*{7}{>{\centering\arraybackslash}m{0.13\linewidth}|}}
      \hline \textbf{Bits} & \textbf{CR3} & \textbf{PDPTE} & \textbf{PDE: 2MB} & \textbf{PDE: PTE} & \textbf{PTE: 4KB} & \textbf{Not Present} \\ \hline
      63 & \multirow{3}*{Ign.} & \multirow{2}*{Rsvd.} & XD & XD & XD & \multirow{15}{*}{Ign.} \\ \cline{1-1} \cline{4-6}
      62:M & & & Rsvd. & Rsvd. & Rsvd. & \\ \cline{1-1} \cline{3-6}
      M-1:32 & & \multirow{4}*{PDE*} & \multirow{2}*{Page*} & \multirow{4}*{PT*} & \multirow{4}*{Page*} & \\ \cline{1-2}
      31:21 & \multirow{10}*{PDPT*} & & & & & \\ \cline{1-1} \cline{4-4}
      20:13 & & & Rsvd. & & & \\ \cline{1-1} \cline{4-4}
      12 & & & PAT & & & \\ \cline{1-1} \cline{3-6}

      11 & & \multirow{3}*{Ign.} & \multirow{3}*{Ign.} & \multirow{4}*{Ign.} & \multirow{3}*{Ign.} & \\ \cline{1-1}
      10 & & & & & & \\ \cline{1-1}
      9 & & & & & & \\ \cline{1-1} \cline{3-4} \cline{6-6}
      8 & & \multirow{4}*{Rsvd.} & G & & G & \\ \cline{1-1} \cline{4-6}
      7 & & & 1 & 0 & PAT & \\ \cline{1-1} \cline{4-6}
      6 & & & D & Ign. & D & \\ \cline{1-1} \cline{4-6}
      5 & & & A & A & A & \\ \cline{1-6}
      4 & \multirow{5}*{Ign.} & PCD & PCD & PCD & PCD & \\ \cline{1-1} \cline{3-6}
      3 & & PWT & PWT & PWT & PWT & \\ \cline{1-1} \cline{3-6}
      2 & & \multirow{2}*{Rsvd.} & U/S & U/S & U/S & \\ \cline{1-1} \cline{4-6}
      1 & & & R/W & R/W & R/W & \\ \cline{1-1} \cline{3-7}
      0 & & 1 & 1 & 1 & 1 & 0 \\ \hline
    \end{tabular}
  \end{table}

  \cite[Chapter~5.4]{sdm3}
} \bigskip

\layout{
  4/5 Level paging:
  \begin{table}[H]
    \centering
    \begin{tabular}{|*{10}{>{\centering\arraybackslash}p{0.099\linewidth}|}}
      \hline \textbf{Bits} & \textbf{CR3} & \textbf{PML5E: PLM4E} & \textbf{PLM4E: PDPTE} & \textbf{PDPTE: 1GB} & \textbf{PDPTE: PDE} & \textbf{PDE: 2MB} & \textbf{PDE: PTE} & \textbf{PTE: 4KB} & \textbf{Not Present} \\ \hline
      63 & \multirow{4}*{Rsvd.} & XD & XD & XD & XD & XD & XD & XD & \multirow{15}{*}{Ign.} \\ \cline{1-1} \cline{3-9}
      62:59 & & \multirow{2}*{Ign.} & \multirow{2}*{Ign.} & Prot & \multirow{2}*{Ign.} & Prot & \multirow{2}*{Ign.} & Prot & \\ \cline{1-1} \cline{5-5} \cline{7-7} \cline{9-9}
      58:52 & & & & Ign. & & Ign. & & Ign. & \\ \cline{1-1} \cline{3-9}
      51:M & & Rsvd. & Rsvd. & Rsvd. & Rsvd. & Rsvd. & Rsvd. & Rsvd. & \\ \cline{1-9}

      M-1:30 & \multirow{4}{=}{\centering PML4* \\ / \\ PML5*} & \multirow{4}{=}{\centering PML4*} & \multirow{4}{=}{\centering PDPT*} & Page* & \multirow{4}{=}{\centering PD*} & \multirow{2}{=}{\centering Page*} & \multirow{4}{=}{\centering PT*} & \multirow{4}{=}{\centering Page*} & \\ \cline{1-1} \cline{5-5}
      29:21 & & & & \multirow{2}{=}{\centering Rsvd.} & & & & & \\ \cline{1-1} \cline{7-7}
      20:13 & & & & & & Rsvd. & & & \\ \cline {1-1} \cline{5-5} \cline{7-7}
      12 & & & & PAT & & PAT & & & \\ \cline{1-9}

      11 & \multirow{7}*{Ign.} & R & R & R & R & R & R & R & \\ \cline{1-1} \cline{3-9}
      10 & & \multirow{3}*{Ign.} & \multirow{3}*{Ign.} & \multirow{2}*{Ign.} & \multirow{3}*{Ign.} & \multirow{2}*{Ign.} & \multirow{3}*{Ign.} & \multirow{2}*{Ign.} & \\ \cline{1-1}
      9 & & & & & & & & & \\ \cline{1-1} \cline{5-5} \cline{7-7} \cline{9-9}
      8 & & & & G & & G & & G & \\ \cline{1-1} \cline{3-9}
      7 & & Rsvd. & Rsvd. & 1 & 0 & 1 & 0 & PAT & \\ \cline{1-1} \cline{3-9}
      6 & & Ign. & Ign. & D & Ign. & D & Ign. & D & \\ \cline{1-1} \cline{3-9}
      5 & & A & A & A & A & A & A & A & \\ \cline{1-9}
      4 & PCD & PCD & PCD & PCD & PCD & PCD & PCD & PCD & \\ \cline{1-9}
      3 & PWT & PWT & PWT & PWT & PWT & PWT & PWT & PWT & \\ \cline{1-9}
      2 & \multirow{3}*{Ign.} & U/S & U/S & U/S & U/S & U/S & U/S & U/S & \\ \cline{1-1} \cline{3-9}
      1 & & R/W & R/W & R/W & R/W & R/W & R/W & R/W & \\ \cline{1-1} \cline{3-10}
      0 & & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 0 \\ \hline
    \end{tabular}
  \end{table}

  \cite[Chapter~5.5]{sdm3}
} \bigskip

\setlength{\tabcolsep}{6pt}

\layout{
  Description of many of the bits/labels used in the page entries:
  \begin{table}[H]
    \centering
    \begin{tabular}{|c|l|l|}
      \hline \textbf{Label} & \textbf{Name} & \textbf{Description} \\ \hline
      P (Bit 0) & Present & If 1 it maps to the next structure else it raises a \#PF \\
      R/W & Read/Write & If 0, writes are not allowed \\
      U/S & User/Supervisor & If 0, user-mode accesses are not allowed \\
      PWT & Page-level Write-Through & Indirectly determines the memory type used to access the page \\
      PCD & Page-level Cache Disable & Indirectly determines the memory type used to access the page \\
      A & Accessed & Indicates whether the software has accessed the page \\
      D & Dirty & Indicates whether the software has written to the page \\
      PAT & Page Attribute Table & Indirectly determines the memory type used to access the page \\
      G & Global & If 1 and CR4.PGE = 1, the translation is global \\
      R & Restart & If 1 and HLAT paging, restart using ordinary paging \\
      Prot & Protection key & If CR4.PKE = 1 or CR4.PKS = 1, the page's access rights, else ignored \\
      XD & Execute Disable & If IA32\_EFER.NXE = 1, Disabled instruction fetches \\
      Ign. & Ignored & Bits are not used \\
      Rsvd. & Reserved & Bits must be 0 \\ \hline
    \end{tabular}
  \end{table}

  \cite[Chapter~5.3]{sdm3} \cite[Chapter~5.4]{sdm3} \cite[Chapter~5.5]{sdm3}
}

% TODO: ??
% 5.6 - Access rights
% 5.7 - page fault exceptions

\layout{
  \subsubsection{Translation Lookaside Buffer}

  The Translation Lookaside Buffer (TLB) is a specialised cache which stores linear address translations, they contain entries that map page numbers to page frames.
  \cite[Chapter~5.10.2]{sdm3}
} \bigskip

\layout{
  Each entry in a TLB is an individual translation;
  each entry is referenced by a page number and stores: physical address of the page frame, access rights, attributes (dirty flag, memory type).
  Other information may also be stored and special-use TLBs may not store all this information (eg instruction TLB would not store the dirty flag).
  TLB entries are associated with the current PCID when they are created.
  \cite[Chapter~5.10.2.2]{sdm3}
} \bigskip

\layout{
  The processor may cache translations that were not used (eg due to speculative execution),
  furthermore some processors may cache a large page (one greater than 4-KiB) as multiple 4-KiB TLB entries.
  \cite[Chapter~5.10.2.3]{sdm3}
} \bigskip

\layout{
  In addition to TLBs there are paging structure caches.
  There is a cache specific to each paging structure, each entry in these caches contains information from the paging structure (eg its physical address and flags) and the bits from the linear address used to identify the entry.
  \cite[Chapter~5.10.3]{sdm3}
} \bigskip

\layout{
  To ensure the paging caches contain up to date information, the OS must invalidate any cached entries when they are updated.
  \cite[Chapter~5.10.4]{sdm3}
} \bigskip

\layout{
  In multiprocessor systems the OS may be required to invalidate the cache in other processors, this requires the OS to perform a TLB shootdown; a shootdown can be achieved by using an interprocessor interrupt (IPI) to interrupt all processors and use the interrupt handler to invalidate the cache.
  \cite[Chapter~5.10.5]{sdm3}
}

\layout{
  \subsection{Virtual Memory}

  Virtual memory is a method of extending main memory by temporarily moving sections of it into secondary storage then moving them back when needed.
  While it is possible to implement virtual memory through segmentation (by moving entire segments in and out of secondary storage), it is most commonly implemented through paging.
} \bigskip

\layout{
  If the OS runs out of memory, it may move several (rarely used) pages into secondary storage (often into a swap file or swap partition), and clearing the present flag for its corresponding page entry.
  If a process attempts to access one of these pages, the page walk will encounter a non-present entry and a \#PF exception will be generated.
  The exception handler will be called and will move the page back into main memory, updating the page entry to the new physical address and setting the present flag.
  When the exception handler returns, the instruction that generated the exception is restarted.

  \bigskip \cite[Chapter~5.12]{sdm3}
}

\section{Instructions}

This section contains many of the x86 instructions and their use, it is not an exhaustive list and is a simplification of how many instructions work (by not including many of the potential exceptions which may be raised, not including how the instructions are encoded in machine code etc, or the flags set/cleared), however it should be sufficient enough to create simple programs.
The format of the instruction is \texttt{instruction op1, op2, op3} eg \texttt{'MOV RAX, 10'}.
Each instruction has different limitations to what type an operand can be, these types are

\begin{table}[H]
  \centering
  \begin{tabular}{|c|c|l|}
    \hline \textbf{Symbol} & \textbf{Name} & \textbf{Description} \\ \hline
    r & General purpose register & A register such as eax \\
    m & Memory address & A memory address such as \texttt{[eax + 4*ebx + 4]} \\
    i & Immediate & A constant value such as 80h or a pointer to a label \\
    s & Segment Register & A register such as ES \\
    d & Debug Register & A register such as DR0 \\
    c & Control Register & A register such as CR3 \\ \hline
  \end{tabular}
\end{table}

The types of an operand or identical instructions with different names will be separated by \texttt{(|)}, the size in bits may also be appended to the type and (') may be used to exclude a single size (eg \texttt{i8|i16} means an 8 or 16 bit immediate value and \texttt{r'8} means a 16, 32 or 64 bit register).
Some instructions have more than one row of operands, each row shows a different combination of operand types. Most instructions should only operate on a single word size (eg 64-bit), if an instruction is used with different sized operands then it may cause an assembler error, cause an invalid opcode exception, truncate the source or bit extend it (either arithmetically or logically).
Furthermore many of these instructions set/clear flags, for example the OF (overflow flag) is set on signed overflow, CF (carry flag) is set on unsigned overflow and ZF (zero flag) is set when the result is zero.
Some instructions can be prefixed with LOCK, this makes the instruction atomic and will give the thread exclusive access to the cache line it is accessing; LOCK can only be used when the destination is a memory location and the instruction supports it else it will generate a \#UD exception.

% TODO: REGISTERS SUCH AS DX:AX, words like qword ptr.
% TODO: (arithmetic) xor, (logical) cmp test, CACHE LINE
% LOCK: ADD, ADC, AND, BTC, BTR, BTS, CMPXCHG, CMPXCH8B, CMPXCHG16B, DEC, INC, NEG, NOT, OR, SBB, SUB, XOR, XADD, and XCHG

\newcommand{\insend}{\end{tabular} \setlength{\tabcolsep}{0pt} \\ \hline} % End of final row of operands
\newcommand{\institle}[1]{#1 \\ \hline}
\newcommand{\insnote}[1]{\makecell[tp{\dimexpr\linewidth-10pt\relax}]{#1} \\ \hline}
\newcommand{\instable}{\setlength{\tabcolsep}{6pt} \begin{tabular}[t]{>{\centering\arraybackslash}p{0.555\linewidth}|>{\centering\arraybackslash}p{0.115\linewidth}|>{\centering\arraybackslash}p{0.115\linewidth}|>{\centering\arraybackslash}p{0.115\linewidth}}}
\newcommand{\inslineb}{\\ & \multicolumn{3}{c}{}}
\newcommand{\inslinea}{\\ \cline{2-4} & \multicolumn{3}{c}{}}
\newcommand{\insmid}{\\ \cline{2-4} &} % End of all other rows of operands
\newcommand{\insop}{\texttt{-} & \texttt{-} & \texttt{-}} % 0 Operands
\newcommand{\insopx}[1]{\texttt{#1} & \texttt{-} & \texttt{-}} % 1 Operand
\newcommand{\insopxx}[2]{\texttt{#1} & \texttt{#2} & \texttt{-}} % 2 Operands
\newcommand{\insopxxx}[3]{\texttt{#1} & \texttt{#2} & \texttt{#3}} % 3 Operands
\newcommand{\insname}[5]{\instable \multirow[t]{#1}{\linewidth}{\makecell[tp{\linewidth}]{\makecell[c]{\texttt{#2} - #3 \cite[Page~#4]{sdm2}} \\ #5}} &}

\setlength{\tabcolsep}{0pt}
\begin{longtable}{|>{\centering\arraybackslash}p{\dimexpr\linewidth-1pt\relax}|}
  \hline \instable \textbf{Instruction} & \textbf{OP 1} & \textbf{OP 2} & \textbf{OP 3} \insend

  \institle{Memory}

  \insname{7}{MOV}{Move}{4-35}{Copies the value from OP2 into OP1}
    \insopxx{r}{r|m|i} \insmid
    \insopxx{m}{r|i'64} \insmid
    \insopxx{d}{r(32|64)} \insmid
    \insopxx{c'CR8}{r32} \insmid
    \insopxx{c}{r64} \insmid
    \insopxx{s}{(r|m)16} \insmid
    \insopxx{(r|m)16}{s} \insend
  \insname{3}{LEA}{Load Effective Address}{3-603}{Stores the address from OP2 into OP1 but unlike MOV it does not dereference the address}
    \insopxx{r'8}{m} \inslinea \inslineb \insend
  \insname{4}{[LOCK] CMPXCHG}{Compare and Exchange}{3-212}{Compares \texttt{AL|AX|EAX|RAX} with OP1 (dst), if equal OP2 (src) is written into OP1 and ZF is set else OP1 is written into the accumulator and ZF is cleared}
    \insopxx{r|m}{r} \inslinea \inslineb \inslineb \insend
  \insname{2}{[LOCK] XCHG}{Exchange}{6-32}{Swaps (exchanges) the values held in OP1 and OP2}
    \insopxx{r}{r|m} \insmid
    \insopxx{r|m}{r} \insend
  \insname{2}{IN}{Input From Port}{3-512}{Copies the value from the I/O port specified by OP2 into OP1}
    \insopxx{al|ax|eax}{dx|i8} \inslinea \insend
  \insname{2}{OUT}{Output To Port}{4-178}{Copies the value from OP2 into the I/O port specified by OP1}
    \insopxx{dx|i8}{al|ax|eax} \inslinea \insend

  \institle{Control}

  \insname{2}{UD}{Undefined Instruction}{4-730}{Causes an invalid opcode exception}
    \insop \inslinea \insend
  \insname{4}{NOP}{No Operation}{4-167}{Performs no operation, the first encoding is a single byte NOP and is an alias for XCHG (E)AX, (E)AX, and the second encoding is a multibyte NOP and may not be supported on all processors}
    \insop \insmid
    \insopx{r|m} \inslinea \inslineb \insend
  \insname{3}{CMP}{Compare Two Operands}{3-179}{Compares OP1 with OP2, it does this by subtracting OP2 from OP1, setting the relevant flag and discarding the result}
    \insopxx{r}{r|m} \insmid
    \insopxx{r|m}{r} \insmid
    \insopxx{r|m}{i'64} \insend
  \insname{3}{TEST}{Logical Compare}{4-714}{Performs a bitwise logical AND on OP1 and OP2, sets the relevant flags and discards the result}
    \insopxx{r|m}{r} \insmid
    \insopxx{r|m}{i'64} \inslinea \insend
  \insname{2}{Jcc}{Jump If Condition Is Met}{3-555}{Jumps if the condition is met}
    \insopx{i'64} \inslinea \insend
  \insname{9}{JMP}{Jump}{3-560}{Jumps to the location in code referenced by OP1. The first encodings are relative signed offsets to the current location in code (i8 is a short jump (type of near jump)), the rest are near jumps (jumps within the current code segment). The second encoding are near jumps to absolute addresses. The third encoding is an immediate of a 16-bit segment selector and an absolute offset into it and performs a far jump into another code segment. The fourth encoding is a memory location that points to a 16-bit segment selector followed by an absolute offset and performs a far jump}
    \insopx{i'64} \insmid
    \insopx{(r|m)'8} \insmid
    \insopx{i16:(16|32)} \insmid
    \insopx{m16:('8)} \inslinea \inslineb \inslineb \inslineb \inslineb \insend

  \institle{Procedures}

  \insname{4}{CALL}{Call Procedure}{3-139}{Similar to \texttt{JUMP}. Additionally, in a near call the current instruction pointer is pushed to the stack. In a far call, both the CS and instruction pointer are pushed to the stack}
    \insopx{i(16|32)} \insmid
    \insopx{(r|m)'8} \insmid
    \insopx{i16:(16|32)} \insmid
    \insopx{m16:('8)} \insend
  \insname{5}{RET}{Return From Procedure}{4-567}{Returns control to the return address located at the top of the stack (usually placed there by a CALL). OP1 specifies the number of bytes to be released from the stack after the return address is popped (default is 0)}
    \insop \insmid
    \insopx{i16} \inslinea \inslineb \inslineb \insend
  \insname{4}{IRET|IRETD|IRETQ}{Interrupt Return}{3-546}{Returns control to the interrupted program from the interrupt handler. It pops off the data pushed to the stack during the interrupt such as EFLAGS, Instruction Pointer and CS}
    \insop \inslinea \inslineb \inslineb \insend
  % TODO: HLT, fences, descriptors, TLB, MSRs, CLI, ENTER, CPUID
  \insname{2}{INT}{Call to Interrupt Procedure}{3-524}{Generates a call to the interrupt handler specified by OP1}
    \insopx{i8} \inslinea \insend

  \institle{Syscalls}
  \insnote{These instructions are used for syscalls (both calling and returning), SYSCALL and SYSRET are better supported on 64-bit CPUs and operating systems. All these instructions set the CS and SS segment selector registers to values defined in MSRs, and their descriptor caches to fixed values with base 0, limit 0xFFFFF, and flags as defined below (in 3 hexadecimal nibbles where the granularity flag is the most significant bit and the type is the last nibble). It is the responsibility of the OS to ensure that the fixed values of the descriptors match the descriptors referenced by the selectors}

  % TODO: if not in 64-bit mode or EFER_EFER.SCE is not set then #UD
  \insname{9}{SYSCALL}{Fast System Call}{4-696}{Invokes the OS system-call handler at privilege level 0, it does so by loading RIP from IA32\_LSTAR MSR and saves the address of the instruction following SYSCALL into RCX. Furthermore RFLAGS is stored into R11, RFLAGS is then masked with NOT(IA32\_FMASK MSR) (clears every bit in RFLAGS which is set in the MSR). The CS selector is loaded with IA32\_STAR[47:32] AND 0xFFFC, and SS is loaded with IA32\_STAR[47:32] + 8. The flags loaded into the CS and SS segment descriptors are 0xA9B and 0xC93 respectively}
    \insop \inslinea \inslineb \inslineb \inslineb \inslineb \inslineb \inslineb \inslineb \insend
  \insname{10}{SYSRET}{Return from Fast System Call}{4-705}{Returns from an OS system-call handler to user code at privilege level 3, it is a companion instruction to SYSCALL. It loads RIP from RCX and RFLAGS from (R11 AND 0x3C7FD7H) OR 2. If the operand size is 64-bit it remains in 64-bit mode else it enters compatibility mode and only the low 32 bits of the registers are loaded. The CS selector is loaded with (IA32\_STAR[63:48] + 16) OR 3 if the operand size is 64-bit else it is loaded with IA32\_STAR[63:48] OR 3, and SS is loaded with (IA32\_STAR[63:48] + 8) OR 3. The flags loaded into the CS segment descriptor are 0xAFB if the operand size is 64-bit and 0xCFB if not and 0xCF3 for the SS segment descriptor}
    \insop \inslinea \inslineb \inslineb \inslineb \inslineb \inslineb \inslineb \inslineb \inslineb \insend
  \insname{10}{SYSENTER}{Fast System Call}{4-699}{Executes a fast call to a level 0 system procedure or routine. In IA-32e mode, it transitions the processor to 64-bit mode, otherwise it remains in protected mode.It loads RIP (or EIP) from 63:0 (or 31:0) of IA32\_SYSENTER\_EIP, and RSP (or ESP) from 63:0 (or 31:0) of IA32\_SYSENTER\_ESP. The CS selector is loaded with IA32\_SYSENTER\_CS[15:0] AND 0xFFFC, and SS is loaded with the value stored in CS + 8. The flags loaded into the CS segment descriptor are 0xA9B in IA-32e mode and 0xC9B if not and 0xC93 for the SS segment descriptor. Since no return information is saved, it is the responsibility of the caller and the OS to manage the return information}
    \insop \inslinea \inslineb \inslineb \inslineb \inslineb \inslineb \inslineb \inslineb \inslineb \insend
  \insname{12}{SYSEXIT}{Fast Return from Fast System Call}{4-702}{Returns to privilege level 3 user code, it is a companion instruction to SYSENTER. If the operand size is 64-bit it remains in 64-bit mode, else it enters compatibility mode or remains in protected mode. RDX (or EDX) is loaded into RIP (or EIP) if the return is not to 64-bit mode then only bits 31:0 are loaded, and RCX (or ECX) is loaded into RSP (or ESP) if the return is not to 64-bit mode then only bits 31:0 are loaded. The CS selector is loaded with (IA32\_SYSENTER\_CS[15:0] + 32) OR 3 if the operand size is 64-bit else it is loaded with (IA32\_SYSENTER\_CS[15:0] + 16) OR 3, and SS is loaded with the value stored in CS + 8. The flags loaded into the CS segment descriptor are 0xAFB if the operand size is 64-bit and 0xCFB if not and 0xCF3 for the SS segment descriptor}
    \insop \inslinea \inslineb \inslineb \inslineb \inslineb \inslineb \inslineb \inslineb \inslineb \inslineb \inslineb \insend

  \institle{Stack}

  % TODO: check this
  \insname{9}{PUSH}{Push data onto the stack}{4-520}{Decrements the Stack Pointer by the operand size, then moves OP1 to the memory location pointed to by the it. It is recommended to only push values the same size as the pointer size to prevent a misaligned stack. Values less than the operand size are sign-extended to the operand size however segment registers are zero-extended. The operand size can be overridden by prefixes however its based on the current stack segment D flag. In long mode, 32-bit registers and the CS, DS, ES and SS registers cannot be encoded}
    \insopx{(r|m)'8} \insmid
    \insopx{i'64} \insmid
    \insopx{s} \inslinea \inslineb \inslineb \inslineb \inslineb \inslineb \insend
  \insname{4}{POP}{Pop data off the stack}{4-396}{Moves the value pointed to by the Stack Pointer to OP1 then increments it by the operand size. Uses the same rules as PUSH}
    \insopx{(r|m)'8} \insmid
    \insopx{s} \inslinea \insend

  \institle{Arithmetic}

  \insname{3}{[LOCK] ADD}{Add}{3-32}{Adds OP2 to OP1. If OP2 is an \texttt{immediate} value and its word size is less than OP1 then it is sign-extended}
    \insopxx{r}{r|m} \insmid
    \insopxx{r|m}{r} \insmid
    \insopxx{r|m}{i'64} \insend
  \insname{2}{[LOCK] INC}{Increment}{3-514}{Increments the value in OP1 by 1}
    \insopx{r|m} \inslinea \insend
  \insname{3}{[LOCK] ADC}{Add With Carry}{3-27}{Adds (OP2 + CF) to OP1. If OP2 is an \texttt{immediate} value and its word size is less than OP1 then it is sign-extended}
    \insopxx{r}{r|m} \insmid
    \insopxx{r|m}{r} \insmid
    \insopxx{r|m}{i'64} \insend
  \insname{3}{[LOCK] SUB}{Subtract}{4-682}{Subtracts OP2 from OP1. If OP2 is an \texttt{immediate} value and its word size is less than OP1 then it is sign-extended}
    \insopxx{r}{r|m} \insmid
    \insopxx{r|m}{r} \insmid
    \insopxx{r|m}{i'64} \insend
  \insname{2}{[LOCK] DEC}{Decrement}{3-326}{Decrements the value in OP1 by 1}
    \insopx{r|m} \inslinea \insend
  \insname{3}{MUL}{Unsigned Multiply}{4-148}{Performs unsigned integer multiplication on \texttt{AL|AX|EAX|RAX} and OP1, and stores the result in \texttt{AX|DX:AX|EDX:EAX|RDX:RAX}}
    \insopx{r|m} \inslinea \inslineb \insend
  \insname{6}{IMUL}{Signed Multipy}{3-508}{Multiplies signed integers. In the first encoding \texttt{AL|AX|EAX|RAX} is multiplied by OP1 and is stored in \texttt{AX|DX:AX|EDX:EAX|RDX:RAX}. In the second OP1 is multiplied by OP2, the result is truncated and stored in OP1. In the third OP1 is multiplied by the product of OP2 and OP3, the result is truncated and stored in OP1}
    \insopx{r|m} \insmid
    \insopxx{r}{r|m} \insmid
    \insopxxx{r}{r|m}{i'64} \inslinea \inslineb \inslineb \insend
  \insname{5}{DIV}{Unsigned Divide}{3-328}{Divides unsigned integers. It divides \texttt{AX|DX:AX|EDX:EAX|RDX:RAX} by OP1, stores the quotient in \texttt{AL|AX|EAX|RAX} and stores the remainder in \texttt{AH|DX|EDX|RDX}. Causes a Divide Error exception if OP1 is 0 or the quotient is larger than its destination register}
    \insopx{r|m} \inslinea \inslineb \inslineb \inslineb \insend
  \insname{5}{IDIV}{Signed Divide}{3-505}{Divides signed integers. It divides \texttt{AX|DX:AX|EDX:EAX|RDX:RAX} by OP1, stores the quotient in \texttt{AL|AX|EAX|RAX} and stores the remainder in \texttt{AH|DX|EDX|RDX}. Causes a Divide Error exception if OP1 is 0 or the quotient is larger than its destination register}
    \insopx{r|m} \inslinea \inslineb \inslineb \inslineb \insend
  \insname{3}{[LOCK] NEG}{Two's Complement Negation}{4-165}{Replaces OP1 with its two's complement (the same as subtracting it from 0)}
    \insopx{r|m} \inslinea \inslineb \insend

  \institle{Logical}

  \insname{2}{BT}{Bit Test}{3-130}{Stores the OP2 (mod OP1 size) bit in OP1 into CF}
    \insopxx{r|m}{r|i8} \inslinea \insend
  \insname{3}{[LOCK] BTC}{Bit Test and Complement}{3-132}{Stores the OP2 (mod OP1 size) bit in OP1 into CF and complements the bit in OP2}
    \insopxx{r|m}{r|i8} \inslinea \inslineb \insend
  \insname{3}{[LOCK] BTR}{Bit Test and Reset}{3-134}{Stores the OP2 (mod OP1 size) bit in OP1 into CF and resets the bit (to 0) in OP2}
    \insopxx{r|m}{r|i8} \inslinea \inslineb \insend
  \insname{3}{[LOCK] BTS}{Bit Test and Set}{3-136}{Stores the OP2 (mod OP1 size) bit in OP1 into CF and sets the bit (to 1) in OP2}
    \insopxx{r|m}{r|i8} \inslinea \inslineb \insend
  \insname{3}{[LOCK] XOR}{Logical Exclusive OR}{6-40}{Performs a bitwise XOR on OP1 and OP2 and stores the result in OP1}
    \insopxx{r|m}{i'64} \insmid
    \insopxx{r|m}{r} \insmid
    \insopxx{r}{r|m} \insend
  \insname{3}{[LOCK] AND}{Logical AND}{3-75}{Performs a bitwise AND on OP1 and OP2 and stores the result in OP1}
    \insopxx{r|m}{i'64} \insmid
    \insopxx{r|m}{r} \insmid
    \insopxx{r}{r|m} \insend
  \insname{3}{[LOCK] OR}{Logical Inclusive OR}{4-170}{Performs a bitwise OR on OP1 and OP2 and stores the result in OP1}
    \insopxx{r|m}{i'64} \insmid
    \insopxx{r|m}{r} \insmid
    \insopxx{r}{r|m} \insend
  \insname{2}{[LOCK] NOT}{One's Complement Negation}{4-168}{Performs a bitwise NOT on OP1 and stores the result in OP1}
    \insopx{r|m} \inslinea \insend
\end{longtable}
\setlength{\tabcolsep}{6pt}

\layout{
  The condition codes (cc) are used in conditional instructions such as Jcc and replace the cc (not all codes are supported by all conditional instructions).
  Different codes check different flags or registers, the codes are described below along with their conditions:

  \begin{table}[H]
    \centering
    \begin{tabular}{|l|l|l|l|}
      \hline \textbf{Code} & \textbf{Description} & \textbf{Flags} & \textbf{Signed-ness} \\ \hline
      O & Overflow & OF = 1 & \\ \hline
      NO & Not Overflow & OF = 0 & \\ \hline
      S & Sign & SF = 1 & \\ \hline
      NS & Not Sign & SF = 0 & \\ \hline
      E & Equal & \multirow{2}*{ZF = 1} & \\ \cline{1-2}
      Z & Zero & & \\ \hline
      NE & Not Equal & \multirow{2}*{ZF = 0} & \\ \cline{1-2}
      NZ & Not Zero & & \\ \hline
      B & Below & \multirow{3}*{CF = 1} & \multirow{3}*{Unsigned} \\ \cline{1-2}
      NAE & Not Above or Equal & & \\ \cline{1-2}
      C & Carry & & \\ \hline
      NB & Not Below & \multirow{3}*{CF = 0} & \multirow{3}*{Unsigned} \\ \cline{1-2}
      AE & Above or Equal & & \\ \cline{1-2}
      NC & Not Carry & & \\ \hline
      BE & Below or Equal & \multirow{2}*{CF = 1 or ZF = 1} & \multirow{2}*{Unsigned} \\ \cline{1-2}
      NA & Not Above & & \\ \hline
      A & Above & \multirow{2}*{CF = 0 and ZF = 0} & \multirow{2}*{Unsigned} \\ \cline{1-2}
      NBE & Not Below or Equal & & \\ \hline
      L & Less & \multirow{2}*{SF != OF} & \multirow{2}*{Signed} \\ \cline{1-2}
      NGE & Not Greater or Equal & & \\ \hline
      GE & Greater or Equal & \multirow{2}*{SF = OF} & \multirow{2}*{Signed} \\ \cline{1-2}
      NL & Not Less & & \\ \hline
      LE & Less or Equal & \multirow{2}*{ZF = 1 or SF != OF} & \multirow{2}*{Signed} \\ \cline{1-2}
      NG & Not Greater & & \\ \hline
      G & Greater & \multirow{2}*{ZF = 0 and SF = OF} & \multirow{2}*{Signed} \\ \cline{1-2}
      NLE & Not Less or Equal & & \\ \hline
      P & Parity & \multirow{2}*{PF = 1} & \\ \cline{1-2}
      PE & Parity Even & & \\ \hline
      NP & Not Parity & \multirow{2}*{PF = 0} & \\ \cline{1-2}
      PO & Parity Odd & & \\ \hline
      CXZ & CX = 0 & & \\ \cline{1-2}
      ECXZ & ECX = 0 & & \\ \cline{1-2}
      RCXZ & RCX = 0 & & \\ \hline
    \end{tabular}
  \end{table}

  \cite{cc-table}
}

% TODO: Finish this section a bit better
% In long mode, to access many of its features the REX prefix has to be placed before an instruction.
% REX must be encoded to access
% 64-bit operands (if the instruction does not default to it)
% an extended register (R8-15, XMM8-15, YMM8-15, CR8-15, DR8-15)
% Uniform byte register (SPL, BPL, SIL, DIL)
% it cannot be encoded when using a high byte register (AH, BH, CH, DH)
% default 64-bit instructions in long mode are:
% CALL (near), ENTER, LEAVE, LGDT, LIDT, LLDT, LOOP, LOOPcc, LTR, MOV CR(n), MOV DR(n), POP r|m, POP r, POP FS, POP GS, POPFQ, PUSH i8, PUSH i32, PUSH r|m, PUSH r, PUSH FS, PUSH GS, PUSHFQ, RET (near)
% Almost every assembler automatically uses the necessary REX prefixes when required so it is not needed to add
% It can be done by using db followed by the required byte in the line above the instruction
% Sometimes REX can cause AH,BH,CH,DH to be interpreted as SPL, BPL, SIL, DIL

% TODO: db, equ, \$ etc

\layout{
  \section{Interrupts}

  Interrupts are signals from devices to the processor causing it to immediately pause execution and run code to handle the interrupt.
  When an interrupt occurs, the processor locates its entry in an OS provided lookup table called the Interrupt Descriptor Table (IDT) and can have up to 256 entries;
  once the processor locates the entry, it jumps to the code to handle the interrupt which is called the Interrupt Service Routine (ISR) or interrupt handler.
  The processor provides functionality to save processor state, if the ISR makes use of this then the interrupted process can be resumed without loss of continuity.
}

\layout{
  There are three types of interrupts:
  \begin{enumerate}
    \item Exceptions: Generated internally by the processor and are used to alert the kernel of an event which requires its attention such as a Double Fault, Page Fault, General Protection Fault, etc
    \item Interrupt Request (IRQ) or Hardware Interrupt: Generated externally by devices on the chipset
    \item Software Interrupt: Generated by software in order to get the kernels attention, often used for syscalls. The \texttt{INT} instruction is used to generate an interrupt
  \end{enumerate}

  \bigskip \cite{osdev-int}
} \bigskip

\layout{
  Each interrupt is generated as an interrupt vector which contains the interrupt vector number.
  Each interrupt condition which requires special handling by the processor is assigned a unique identification number (vector number).
  The first 32 vectors numbers are reserved by Intel for exceptions.

  \bigskip \cite[Chapter~7.2]{sdm3}
} \bigskip

\layout {
  There are three classifications of interrupt:
  \begin{enumerate}
    \item Faults - Generally correctable, allows the program to restart with no loss of continuity.
      Processor restores state prior to faulting instruction which is restarted when returned.
    \item Traps - Reported immediately following the execution of a trapping instruction, allows the program to restart with no loss of continuity.
    \item Aborts - Does not always report the precise location of the instruction, does not allow the restart of the program.
  \end{enumerate}

  \cite[Chapter~7.5]{sdm3}
} \bigskip

\layout{
  Nonmaskable Interrupts (NMIs) are interrupts that cannot be masked (eg by EFLAGS.IF), furthermore subsequent NMIs are blocked until the current NMI handler returns.

  \bigskip \cite[Chapter~7.7]{sdm3}
}

\layout{
  \subsection{Sources of interrupts}

  Interrupts can be generated in two ways: hardware generated and software generated.

  \bigskip \cite[Chapter~7.3]{sdm3}
} \bigskip

\layout{
  The \texttt{INT} instruction is used to allow software to generate an interrupt by providing the interrupt vector number as an operand.
  The instruction is able to generate interrupts for the architecturally defined interrupts, however it does not push the error code onto the stack which would cause the ISR to return to an incorrect location.
  Furthermore calling \texttt{INT 2} will call the NMI handler without the processors NMI-handling hardware, furthermore software interrupts cannot be masked with EFLAGS.IF.

  \bigskip \cite[Chapter~7.3.3]{sdm3} \cite[Chapter~7.4.2]{sdm3}
} \bigskip

\layout{
  Hardware interrupts generated by the processor:
  \begin{table}[H]
    \centering
    \begin{tabular}{|c|c|c|c|l|}
      \hline \textbf{Vector} & \textbf{Mnemonic} & \textbf{Type} & \textbf{Error Code} & \textbf{Description} \\ \hline
      0 & \#DE & Fault & No & Divide Error Exception \\
      1 & \#DB & Fault/Trap & No &Debug Exception \\
      2 & - & Interrupt & No & NMI Interrupt \\
      3 & \#BP & Trap & No & Breakpoint Exception \\
      4 & \#OF & Trap & No & Overflow Exception \\
      5 & \#BR & Fault & No & BOUND Range Exceeded Exception \\
      6 & \#UD & Fault & No & Invalid Opcode Exception \\
      7 & \#NM & Fault & No & Device Not Available Exception \\
      8 & \#DF & Abort & Yes (zero) &  Double Fault Exception \\
      9 & - & Fault & No & Coprocessor Segment Overrun \\
      10 & \#TS & Fault & Yes & Invalid TSS Exception \\
      11 & \#NP & Fault & Yes & Segment Not Present \\
      12 & \#SS & Fault & Yes & Stack Fault Exception \\
      13 & \#GP & Fault & Yes & General Protection Exception \\
      14 & \#PF & Fault & Yes & Page-Fault Exception \\
      16 & \#MF & Fault & No & x87 FPU Floating-Point Error \\
      17 & \#AC & Fault & Yes & Alignment Check Exception \\
      18 & \#MC & Abort & No & Machine-Check Exception \\
      19 & \#XM & Fault & No & SIMD Floating-Point Exception \\
      20 & \#VE & Fault & No & Virtualisation Exception \\
      21 & \#CP & Fault & Yes & Control Protection Exception \\ \hline
    \end{tabular}
  \end{table}

  Furthermore any vectors not listed in the range 0 to 31 are reserved and should not be used, all vectors in the range 32 to 255 are user defined interrupts.

  \bigskip \cite[Chapter~7.3.1]{sdm3}
}

\layout{
  \subsection{ISRs}

  Interrupt Service Routines (ISRs) are code that handle interrupts, they generally should preserve state of the processor by pushing any registers to the stack before using them and popping them off before returning to the previous process by using the \texttt{IRET} instruction.
} \bigskip

\layout{
  In protected mode, if there was no CPL change when calling the interrupt then the stack uses the interrupted program's stack,
  it pushes: EFLAGS, CS, EIP, then the Error Code (if there was an error code).
  If there was a CPL change then it switches the stack (SS and ESP for the handler are obtained from the TSS) and pushes to the new one: SS, ESP, EFLAGS, CS, EIP, then the Error Code (if there was an error code), furthermore in this case \texttt{IRET also restores the original stack}.
  % TODO: shadow stack?

  \bigskip \cite[Chapter~7.12.1]{sdm3}
} \bigskip

\layout {
  % TODO: finish (64 bit stack contents + stack switches)
  7.14

  In IA-32e mode:
  Handlers are 64-bit
  Interrupt-stack pushes is fixed at 64 bits (zero extended stores).
  SS:RSP always pushed.
  SS is set to null on CPL change.
  IRET behaviour changes
  new stack switch/shadow stack
  Alignment of interrupt stack frame is different


} \bigskip

\layout{
  Format of the error code:
  \begin{table}[H]
    \centering
    \begin{tabular}{|c|c|m{0.7\linewidth}|}
      \hline \textbf{Bits} & \textbf{Label} & \textbf{Description} \\ \hline
      31:16 & Rsvd. & Must be 0 \\
      15:3 & Selector & Segment Selector Index \\
      2 & TI & When set it indicated the selector references the LDT, otherwise it references the LDT (Only used when IDT is clear) \\
      1 & IDT & When set it indicates the selector references the IDT, otherwise it references the GDT/LDT \\
      0 & EXT & When set it occured due to an external event, if cleared it was caused by an \texttt{INT} instruction \\ \hline
    \end{tabular}
  \end{table}

  % TODO: finish
  Format is different for \#PF and \#CF.
  Reserved portion may not be pushed as to align the stack to 32-bits.
  IRET does not pop off error code so ISR must do it.

  \bigskip \cite[Chapter~7.13]{sdm3}
}

\layout{
  \subsection{PIC}

  IBM-PC compatible environments contain a PIC (Programmable Interrupt Controller).
  The PIC has been replaced by the APIC which should be used instead.
} \bigskip

\layout{
  Most systems contain two PICs, each with 8 inputs and a single output; the slave PIC's output is connected to the master PIC's third input (input \#2).
  When the slave PIC attempts to interrupt the processor, it sends it to the master PIC which then interrupts the processor, this means that IRQ \#2 cannot occur.
  Each PIC has an 8-bit Interrupt Mask Register (IMR), when a bit is set the PIC ignores the request on that line.
} \bigskip

\layout{
  External devices send a signal to the PIC when they need to interrupt the processor (without the PIC the processor  would have to poll these devices).
  The PIC translates the IRQ number (from the interrupt line the signal was on) into an interrupt vector, where it then interrupts the processor.
  On system start up, IRQs 0:7 are set to interrupts 0x08:0x0F and IRQs 8:15 are set to interrupts 0x70:0x77 by the BIOS.
  IQRs are handled by priority level: 0, 1, 2, 8, 9, 10, 11, 12, 13, 14, 15, 3, 4, 5, 6, 7.
  The ISR handles the interrupt by communicating with the device via the \texttt{in} and \texttt{out} instructions, once the ISR has finished it should inform the PIC that it can send new interrupts.
  In order to inform the master PIC, 20h should be sent to port 20h;
  to inform the slave PIC, 20h should be send to port A0h and port 20h.
} \bigskip

\layout{
  The PICs can be remapped so IRQs are mapped to different interrupt vectors (useful since the default interrupts for several IRQs are reserved), however they can only be mapped to multiples of 8.
  To remap/initialise the PIC, send it 0x11 to begin initialisation on its command port, then on the data port send it: the vector offset, then 0x04 if its the master and 0x02 if its the slave, finally send it 0x01 (operate in 8086 mode).
} \bigskip

\layout{
  Standard ISA IRQs:
  \begin{table}[H]
    \centering
    \begin{tabular}{|c|l|}
      \hline \textbf{IRQ} & \textbf{Description} \\ \hline
      0 & Programmable Interrupt Timer Interrupt \\
      1 & Keyboard Interrupt \\
      2 & Cascade (used internally by the PICs, never raised) \\
      3 & COM2 (if enabled) \\
      4 & COM1 (if enabled) \\
      5 & LPT2 (if enabled) \\
      6 & Floppy disk \\
      7 & LPT1 / Unreliable spurious interrupt \\
      8 & CMOS real-time clock (if enabled) \\
      9 & Free for peripherals / legacy SCSI / NIC \\
      10 & Free for peripherals / SCSI / NIC \\
      11 & Free for peripherals / SCSI / NIC \\
      12 & PS2 Mouse \\
      13 & FPU / Coprocessor / Inter-processor \\
      14 & Primary ATA Hard Disk \\
      15 & Secondary ATA Hard Disk \\ \hline
    \end{tabular}
  \end{table}
}

\layout{
  Ports used by PIC:
  \begin{table}[H]
    \centering
    \begin{tabular}{|c|l|}
      \hline \textbf{Port} & \textbf{Description} \\ \hline
      20h & Master PIC control \\
      21h & Master PIC mask/data \\
      A0h & Slave PIC control \\
      A1h & Slave PIC mask/data \\
      60h & Keyboard controller data \\
      64h & Keyboard controller command \\ \hline
    \end{tabular}
  \end{table}
}

\layout{
  Example keyboard ISR: \\
  \texttt{push eax ; Save state} \\
  \texttt{in al, 60h ; Read from keyboard (al contains scancode)} \\
  \texttt{mov al, 20h} \\
  \texttt{out 20h, al ; Acknowledge interrupt} \\
  \texttt{pop eax ; Restore state} \\
  \texttt{iret ; Return from ISR}

  \bigskip \cite{osdev-int} \cite{osdev-pic}
}

\subsection{APIC}

The APIC (Advanced Programmable Interrupt Controller) is the updated standard for the older PIC.
The APIC is used for sophisticated interrupt redirection and for sending interrupts between processors in multiprocessor systems.
When using the APIC the PICs should be disabled, this can be done by sending 0xFF to both PIC mask ports.

\subsection{IPI}

\layout{
  \subsection{IDT}

  The Interrupt Descriptor Table associates each interrupt vector with an interrupt descriptor for the ISR to handle the interrupt.
  The linear base address of the IDT and limit (size of IDT - 1) is contained in the IDT register (IDTR), which can be written to using the \texttt{SIDT} instruction.
  Since there are only 256 interrupt vectors, the IDT should not contain more than 265 descriptors.
  In protected mode each descriptor is 8 bytes and in IA-32e mode each descriptor is 16 bytes.
  When an interrupt occurs the interrupt vector provides a vector number which is used to index into the IDT, and uses the descriptor to call the corresponding ISR handler.

  \bigskip \cite[Chapter~7.10]{sdm3}
} \bigskip

\layout{
  In protected mode there are three types if interrupt descriptor, whilst in IA-32e mode there are only two.

  \begin{enumerate}
    \item Task-gate descriptor (protected mode only)
    \item Interrupt-gate descriptor (both)
    \item Trap-gate descriptor (both)
  \end{enumerate}
}

\layout{
  Task-gate descriptor (protected)
  \begin{table}[H]
    \centering
    \begin{tabular}{|c|c|l|}
      \hline \textbf{Bits} & \textbf{Label} & \textbf{Description} \\ \hline
      63:48 & Rsvd. & Must be 0 \\
      47 & P & Segment Present Flag \\
      46:45 & DPL & Descriptor Privilege Level \\
      44:40 & 00101 & Must be 00101 \\
      39:32 & Rsvd. & Must be 0 \\
      31:16 & CS & TSS Segment Selector \\
      15:0 & Rsvd. & Must be 0 \\ \hline
    \end{tabular}
  \end{table}
}

\layout{
  Interrupt-gate descriptor (protected)
  \begin{table}[H]
    \centering
    \begin{tabular}{|c|c|l|}
      \hline \textbf{Bits} & \textbf{Label} & \textbf{Description} \\ \hline
      63:48 & Offset 31:16 & Bits 31:16 of the address of the ISR \\
      47 & P & Segment Present Flag \\
      46:45 & DPL & Descriptor Privilege Level \\
      44 & 0 & Must be 0 \\
      43 & D & 1 = 32 bit, 0 = 16 bit \\
      42:32 & 11000000000 & Must be 11000000000 \\
      31:16 & CS & Code Segment Selector \\
      15:0 & Offset 15:0 & Bits 15:0 of the address of the ISR \\ \hline
    \end{tabular}
  \end{table}
}

\layout{
  Trap-gate descriptor (protected)
  \begin{table}[H]
    \centering
    \begin{tabular}{|c|c|l|}
      \hline \textbf{Bits} & \textbf{Label} & \textbf{Description} \\ \hline
      63:48 & Offset 31:16 & Bits 31:16 of the address of the ISR \\
      47 & P & Segment Present Flag \\
      46:45 & DPL & Descriptor Privilege Level \\
      44 & 0 & Must be 0 \\
      43 & D & 1 = 32 bit, 0 = 16 bit \\
      42:32 & 11100000000 & Must be 11100000000 \\
      31:16 & CS & Code Segment Selector \\
      15:0 & Offset 15:0 & Bits 15:0 of the address of the ISR \\ \hline
    \end{tabular}
  \end{table}

  \cite[Chapter~7.11]{sdm3}
} \bigskip

\layout{
  64-bit Interrupt/trap gate:
  \begin{table}[H]
    \centering
    \begin{tabular}{|c|c|l|}
      \hline \textbf{Bits} & \textbf{Label} & \textbf{Description} \\ \hline
      127:96 & Rsvd. & Reserved (must be 0) \\
      95:64 & Offset 63:32 & Bits 63:32 of the address of the ISR \\
      63:48 & Offset 31:16 & Bits 31:16 of the address of the ISR \\
      47 & P & Segment Present Flag \\
      46:45 & DPL & Descriptor Privilege Level \\
      44 & 0 & Must be 0 \\
      43:40 & Type & Type of the Descriptor \\
      39:35 & 0 & Must be 0 \\
      34:32 & IST & Interrupt Stack Table \\
      31:16 & CS & Code Segment Selector \\
      15:0 & Offset 15:0 & Bits 15:0 of the address of the ISR \\ \hline
    \end{tabular}
  \end{table}

  \cite[Chapter~7.14]{sdm3}

  % TODO: check the gates
  % TYPE = 1110 (interrupt) or 1111 (trap)
}

\subsection{BIOS interrupts}

% \subsection{UEFI interrupts}

\subsection{Syscalls}

\layout{
  In order for a program to request resources from the OS it has to hand over control of the processor to the OS via a syscall/interrupt.
  32-bit programs use the interrupt \texttt{int 80h}, while 64-bit programs use the instruction \texttt{syscall}.
  The parameters for the syscall are placed in specific registers, the following is a list of registers used and their use in syscalls:

  \begin{table}[H]
    \centering
    \begin{tabular}{|c|c|c|}
      \hline \textbf{i386 Registers} & \textbf{x86-64 Registers} & \textbf{Usage} \\ \hline
      EBX & RDI & Arg1 \\
      ECX & RSI & Arg2 \\
      EDX & RDX & Arg3 \\
      ESI & R10 & Arg4 \\
      EDI & R8 & Arg5 \\
      EBP & R9 & Arg6 \\
      EAX & EAX & Syscall \\
      EAX & RAX & Return \\
      \hline
    \end{tabular}
  \end{table}
}

\layout{
  Furthermore, the values in these registers (and others such as R11 and RCX) may be clobbered (their value changing after the syscall) so it is best to follow the System V ABI when performing syscalls.
  When using the syscall function in C if the syscall had a negative return (error), it is negated and copied to errno, and the function returns -1.

  \bigskip \cite{man-syscall}
} \bigskip

\newcommand{\sysd}[1]{\hline \multicolumn{10}{|p{0.97\linewidth}|}{#1} \\}
\newcommand{\sysxxxxxx}[6]{\sysxxxxx{#1}{#2}{#3}{#4}{#5} \\ 6: #6}
\newcommand{\sysxxxxx}[5]{\sysxxxx{#1}{#2}{#3}{#4} \\ 5: #5}
\newcommand{\sysxxxx}[4]{\sysxxx{#1}{#2}{#3} \\ 4: #4}
\newcommand{\sysxxx}[3]{\sysxx{#1}{#2} \\ 3: #3}
\newcommand{\sysxx}[2]{\sysx{#1} \\ 2: #2}
\newcommand{\sysx}[1]{1: #1}
\newcommand{\sys}{}

\newcommand{\sysentry}[6]{
  \begin{tabular}[t]{>{\centering\arraybackslash}p{0.8\linewidth}|>{\centering\arraybackslash}p{0.2\linewidth}}
    \begin{tabular}[t]{p{\linewidth}}
      \begin{tabular}[t]{*{3}{>{\centering\arraybackslash}p{0.249\linewidth}|}>{\centering\arraybackslash}p{0.249\linewidth}}
        \texttt{#1} & \texttt{#2} & \texttt{#3} & \texttt{#4}
      \end{tabular} \\ \hline
      \setlength{\tabcolsep}{6pt}
      \begin{tabular}[t]{p{0.97\linewidth}}
        #5
      \end{tabular}
      \setlength{\tabcolsep}{0pt}
    \end{tabular} &
    \setlength{\tabcolsep}{6pt}
    \begin{tabular}[t]{p{\dimexpr\linewidth-12pt\relax}}
      #6
    \end{tabular}
    \setlength{\tabcolsep}{0pt}
  \end{tabular} \\ \hline
}

The following table contains the arguments, descriptions, return information, and syscall number (for both i386\cite{syscall-number-32} and x86-64\cite{syscall-number-64}) for various common Linux syscalls:

\setlength{\tabcolsep}{0pt}
\begin{longtable}{|p{0.975\linewidth}|}
  \hline \begin{tabular}[t]{>{\centering\arraybackslash}p{0.8\linewidth}|>{\centering\arraybackslash}p{0.2\linewidth}}
    \begin{tabular}[t]{p{\linewidth}}
      \begin{tabular}[t]{*{3}{>{\centering\arraybackslash}p{0.249\linewidth}|}>{\centering\arraybackslash}p{0.249\linewidth}}
        \textbf{Name} & \textbf{Return} & \textbf{i386} & \textbf{x86-64}
      \end{tabular}
    \end{tabular} & \textbf{Arguments}
  \end{tabular} \\ \hline

  \sysentry{read\cite{man-read}}{length|error}{3}{0}{Attempts to read up to \texttt{count} bytes from the current position of the file referred to by \texttt{fd} into the buffer pointed to by \texttt{buf}. The true number of bytes read is returned (if EOF less bytes would be read) and the position in the file is advanced by this many bytes, otherwise returns an error}{\sysxxx{fd}{buf*}{count}}
  \sysentry{write\cite{man-write}}{length|error}{4}{1}{Writes up to \texttt{count} bytes from \texttt{buf} into the file referred to by \texttt{fd} at the current position of the file, and its position is advanced by the number of bytes written. On success the number of bytes written is returned, otherwise returns an error}{\sysxxx{fd}{buf*}{count}}
  \sysentry{open\cite{man-open}}{fd|error}{5}{2}{Opens the file located at \texttt{path} using the given \texttt{flags}. On success returns the file descriptor (\texttt{fd}), otherwise returns an error. The mode argument is for the permissions of the created file when O\_CREAT or O\_TMPFILE is given as a flag and is otherwise ignored}{\sysxxx{path*}{flags}{mode}}
  \sysentry{close\cite{man-close}}{0|error}{6}{3}{Closes the file descriptor (\texttt{fd}). On success returns 0, otherwise returns an error}{\sysx{fd}}
  \sysentry{fstat\cite{man-stat}}{0|error}{108}{5}{Writes information about the file referred to by \texttt{fd} into the \texttt{struct stat\cite{man-stat-type}} pointed to by \texttt{statbuf}. On success returns 0, otherwise returns an error}{\sysxx{fd}{statbuf*}}
  \sysentry{exit\cite{man-exit}}{noreturn}{1}{60}{Terminates the calling thread, \texttt{status \& 0xFF} is returned to the parent process (if the thread was the last in the process)}{\sysx{status}}
  \sysentry{munmap\cite{man-mmap}}{0|error}{91}{11}{All pages containing part of the specified address range are unmapped (the region is also unmapped when the process is terminated). \texttt{addr} must be a multiple of page size but \texttt{length} does not need to be. On success 0 is returned, otherwise an error is returned (it is not an error if there were no mapped pages within the range)}{\sysxx{addr*}{length}}
  \sysentry{mmap\cite{man-mmap}}{addr*|error}{90}{9}{Creates a new mapping in the virtual address space of the process. The kernel attempts to create a mapping aligned to a page boundary close to the \texttt{addr}, or chooses one itself if \texttt{addr} is NULL. The contents of a file mapping are initialised using \texttt{length} bytes starting at \texttt{offset} (multiple of page size) in the file referred to by \texttt{fd}, the file descriptor can be closed after mmap has returned without invalidating the mapping. \texttt{prot} describes the memory protection of the mapping (must not conflict with the open mode of the file); it is a bitwise OR of READ (1), WRITE(2) and EXEC (4). \texttt{flags} determines if updates to the mapping are visible to other processes with mappings to the same region, and if they are carried through to the underlying file; one important flag is MAP\_ANONYMOUS (0x20) for mappings not backed by files, \texttt{fd} should be -1 and \texttt{offset} should be 0. On success the address of the mapping is returned, otherwise an error is returned}{\sysxxxxxx{addr*}{length}{prot}{flags}{fd}{offset}}
\end{longtable}
\setlength{\tabcolsep}{6pt}

\layout{
  A fd (file descriptor) in Linux is an integer used to refer to a file, when opening a file the kernel returns an integer to the program to refer to the file with (lowest unused integer) and the kernel keeps track of the open file.
  Integers 0, 1 and 2 refer to stdin, stdout and stderr respectively; any negative fd is invalid -1 is usually used to refer to no file\cite{unix-fd}.
  Furthermore any arguments in the table which contain a * such as buf* are pointers.
}

\section{Appendix}
All of the code used/written has been appended below, it includes programs that can run on Linux or baremetal

\subsection{Bash scripts}
These bash scripts were used to run the code.

\noindent \newline 64-bit Linux code (Links together files if given more than one argument)
\lstinputlisting{../run64.sh}
32-bit Linux code
\lstinputlisting{../run32.sh}

\subsection{Hello world (Linux)}
This program writes "Hello World" to stdout then exits without an error.

\noindent \newline 64-bit code:
\lstinputlisting{../print/64.asm}
32-bit code:
\lstinputlisting{../print/32.asm}

\subsection{Difference (Linux)}
This program reads two integers from stdin and writes their difference to stdout

\lstinputlisting{../subtraction/64.asm}

\subsection{Cat (Linux)}
This program performs the cat command (it takes in one argument which is a filename and prints its contents). It uses the second arg (the first being the path) as the filename for the open syscall, it then calls the fstat syscall on the returned file descriptor. It gets the size of the file from fstat and allocates the file to memory (the size allocated is the next multiple of page size of the file size). It then writes the memory to stdout then deallocates the memory, closes the file and exits. It also writes to stderr and exits on critical errors (ones that would cause undefined behaviour) but ignores non-critical errors

\lstinputlisting{../cat/cat.asm}

\subsection{Hello world (Baremetal)}
This program prints "Hello world" to the vga-mode buffer.

\lstinputlisting{../baremetal/main.asm}

\subsection{Kernel (Baremetal)}
This program loads a (very) basic 64-bit kernel, it is capable of booting into a 64-bit environment using identity-mapped memory and can handle PS/2 keyboard interrupts. It starts execution in the 16-bit assembly, which sets up a 32-bit environment and calls \_\_start\_\_ in bootloader.c which sets up the 64-bit environment, returns to the assembly code to enable 64-bit mode then jumps to \_\_start\_\_ in kernel.c.

This bash script is used to compile and run the kernel (using qemu)
\lstinputlisting{../kernel/run.sh}

These files are used in the compilation of the kernel

link/ld32
\lstinputlisting{../kernel/link/ld32}
link/ld64
\lstinputlisting{../kernel/link/ld64}
link/link32
\lstinputlisting{../kernel/link/link32}
link/link32
\lstinputlisting{../kernel/link/link64}

\newcommand{\cfile}[1]{
  #1.c
  \lstinputlisting[language=c]{../kernel/src/#1.c}
}

\newcommand{\chfile}[1]{
  #1.h
  \lstinputlisting[language=c]{../kernel/include/#1.h}
  #1.c
  \lstinputlisting[language=c]{../kernel/src/#1.c}
}

The following files are the c, header and assembly files for the kernel

bootloader.asm
\lstinputlisting{../kernel/bootloader.asm}

\cfile{bootloader}
\chfile{utils}
\cfile{kernel}
\chfile{identity}
\chfile{interrupts}
\chfile{shared\_constants}

\section{References}

\hbadness=10000
\bibliographystyle{unsrt}
\bibliography{refs}

\end{document}
